---
title: "Choosing lambda (low-mismatch case)"
author: "Dat Do"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)

library(susieR)

source("code/SuSiE-relax.R")
source("code/sim_2pop.R")
source("code/plot.R")
```

# Goal

SuSiE-relax does not use the out-of-sample LD matrix \(R_0\) directly. It uses a
stabilized reference
\[
\bar R = (1-\lambda) R_0 + \lambda I,
\]
re-scaled to a correlation matrix. The ridge parameter \(\lambda\) controls how
much \(R_0\) is pulled toward the identity before fitting. This note
investigates whether the choice of \(\lambda\) affects the fine-mapping result
in the **low-mismatch** case, where the reference panel is large enough that
\(R_0\) is already close to the in-sample LD.

The plan is:

1. Simulate one low-mismatch data set (the same settings as the low-mismatch
   case in the SuSiE-relax demonstration).
2. Build \(\bar R\) for a grid of \(\lambda\) values and fit SuSiE-relax to each.
3. Plot the final ELBO of each fit against \(\lambda\).
4. Report the PIP result of the best \(\lambda\) (largest final ELBO) and
   compare it to ordinary susieR fit on \(R_0\).

Because the ELBO is a variational lower bound on \(\log p(x \mid \bar R)\) and
\(x\) is held fixed, comparing the final ELBO across the \(\lambda\) grid is a
reasonable model-selection criterion for \(\lambda\).

```{r helper-functions}
standardize_sim <- function(res) {
  d <- 1 / sqrt(diag(res$R))
  res$v <- d * res$v
  res$R <- cov2cor(res$R)
  res$R0 <- cov2cor(res$R0)
  res
}

make_rbar <- function(R0, lambda = 0.01) {
  Rbar <- (1 - lambda) * R0 + lambda * diag(1, ncol(R0))
  cov2cor(Rbar)
}

make_cs_from_alpha <- function(alpha, R, coverage = 0.95,
                               min_abs_corr = 0.5,
                               min_top_alpha = 1e-4) {
  alpha <- as.matrix(alpha)
  cs <- list()
  purity <- numeric()

  for (ell in seq_len(nrow(alpha))) {
    a <- as.numeric(alpha[ell, ])
    if (max(a) < min_top_alpha) {
      next
    }

    ord <- order(a, decreasing = TRUE)
    k <- which(cumsum(a[ord]) >= coverage)[1]
    vars <- ord[seq_len(k)]
    p <- if (length(vars) == 1) {
      1
    } else {
      R_sub <- abs(R[vars, vars, drop = FALSE])
      min(R_sub[upper.tri(R_sub)])
    }

    if (is.finite(p) && p >= min_abs_corr) {
      nm <- paste0("L", ell)
      cs[[nm]] <- vars
      purity[nm] <- p
    }
  }

  list(cs = cs, purity = purity)
}

count_selected_cs <- function(alpha, R, coverage = 0.95,
                              min_abs_corr = 0.5,
                              min_top_alpha = 1e-4) {
  length(make_cs_from_alpha(
    alpha = alpha,
    R = R,
    coverage = coverage,
    min_abs_corr = min_abs_corr,
    min_top_alpha = min_top_alpha
  )$cs)
}

summarize_fit <- function(label, alpha, pip, R_for_cs, beta_true,
                          nu0 = NA_real_) {
  alpha <- as.matrix(alpha)
  top <- apply(alpha, 1, which.max)
  top_alpha <- apply(alpha, 1, max)
  cs_info <- make_cs_from_alpha(alpha, R = R_for_cs)

  data.frame(
    method = label,
    selected_L = length(cs_info$cs),
    nu0 = nu0,
    gamma_hat = paste(top, collapse = ", "),
    top_alpha = paste(sprintf("%.3f", top_alpha), collapse = ", "),
    causal_pip = paste(sprintf("%.3f", pip[beta_true != 0]), collapse = ", ")
  )
}

plot_pip <- function(fit, R_for_cs, beta_true, main) {
  susie_plot_alpha(
    fit,
    R = R_for_cs,
    b = beta_true,
    main = main,
    coverage = 0.95,
    min_abs_corr = 0.5
  )
  grid()
}
```

# Simulate the low-mismatch data

We use the same simulation settings as the low-mismatch case in the SuSiE-relax
demonstration: a large reference panel (\(N_0 = 2000\)) so that the
out-of-sample LD is close to the GWAS LD. The data set is simulated once and
shared across all \(\lambda\) values.

```{r simulate}
N  <- 20000L
J  <- 1000L
L  <- 5L

res <- sim_2pop(
  chrom = "chr22",
  start = 20e6,
  end = 21e6,
  T_split = 20,
  n_gwas = N,
  n_ref = 2000L,
  J = J,
  h2 = 0.01,
  seed = 6
)
res <- standardize_sim(res)

v         <- res$v
R         <- res$R
R0        <- res$R0
beta_true <- res$beta_true
causal    <- which(beta_true != 0)

causal
```

For reference, we also fit ordinary susieR directly on the out-of-sample LD
\(R_0\). This fit does not depend on \(\lambda\).

```{r fit-susieR}
fit_susie_out <- susie_suff_stat(
  XtX = N * R0,
  Xty = N * v,
  n = N,
  yty = N,
  L = L
)
```

# Fit SuSiE-relax over a grid of lambda

```{r lambda-grid-fit}
lambda_grid <- c(1e-5, 1e-4, 5e-4, 1e-3, 5e-3, 1e-2, 5e-2, 1e-1)

relax_fits <- lapply(lambda_grid, function(lambda) {
  Rbar <- make_rbar(R0, lambda = lambda)
  fit_susie_relax(
    x = v,
    Rbar = Rbar,
    N = N,
    L = L,
    sigma2 = rep(0.3^2, L),
    nu0_init = 2000,
    estimate_nu0 = TRUE,
    warmup_iter = 5,
    max_iter = 1000,
    tol = 1e-7,
    verbose = FALSE,
    nu0_bounds = c(10, 10000),
    nu0_update_interval = 1,
    nu0_tol = 1e-4
  )
})
names(relax_fits) <- sprintf("lambda=%g", lambda_grid)

lambda_summary <- do.call(rbind, lapply(seq_along(lambda_grid), function(i) {
  fit <- relax_fits[[i]]
  data.frame(
    lambda      = lambda_grid[i],
    final_elbo  = fit$elbo[length(fit$elbo)],
    n_iter      = length(fit$elbo),
    nu0         = fit$nu0,
    tau2        = fit$tau2,
    selected_L  = count_selected_cs(fit$alpha, R = make_rbar(R0, lambda_grid[i])),
    causal_pip  = paste(sprintf("%.3f", fit$pip[beta_true != 0]), collapse = ", ")
  )
}))

lambda_summary
```

# Final ELBO against lambda

```{r elbo-vs-lambda}
best_idx    <- which.max(lambda_summary$final_elbo)
best_lambda <- lambda_summary$lambda[best_idx]

plot(
  lambda_summary$lambda, lambda_summary$final_elbo,
  type = "b", lwd = 2, pch = 16, log = "x",
  xlab = expression(lambda), ylab = "Final ELBO",
  main = "Low mismatch: final ELBO vs lambda"
)
points(
  best_lambda, lambda_summary$final_elbo[best_idx],
  col = "red", pch = 1, cex = 2.5, lwd = 2
)
grid()

best_lambda
```

The red circle marks the \(\lambda\) with the largest final ELBO.

# PIP of the best lambda vs susieR on R0

```{r best-summary}
best_fit  <- relax_fits[[best_idx]]
best_Rbar <- make_rbar(R0, lambda = best_lambda)

summary <- rbind(
  summarize_fit(
    label = "susieR with R0",
    alpha = fit_susie_out$alpha,
    pip = fit_susie_out$pip,
    R_for_cs = R0,
    beta_true = beta_true
  ),
  summarize_fit(
    label = sprintf("SuSiE-relax (best lambda = %g)", best_lambda),
    alpha = best_fit$alpha,
    pip = best_fit$pip,
    R_for_cs = best_Rbar,
    beta_true = beta_true,
    nu0 = best_fit$nu0
  )
)

summary
```

```{r best-pip-plots}
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
plot_pip(
  fit_susie_out,
  R0,
  beta_true,
  main = "Low mismatch: susieR with R0"
)
plot_pip(
  best_fit,
  best_Rbar,
  beta_true,
  main = sprintf("SuSiE-relax (lambda = %g)", best_lambda)
)
par(old_par)
```
