---
title: "Choosing lambda (different-ancestry case)"
author: "Dat Do"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)

library(susieR)

source("code/SuSiE-relax.R")
source("code/plot.R")
```

# Goal

SuSiE-relax does not use the out-of-sample LD matrix \(R_0\) directly. It uses a
stabilized reference
\[
\bar R = (1-\lambda) R_0 + \lambda I,
\]
re-scaled to a correlation matrix. The ridge parameter \(\lambda\) controls how
much \(R_0\) is pulled toward the identity before fitting.

This note investigates the choice of \(\lambda\) in a different-ancestry
mismatch case. The GWAS sample is simulated from CHB haplotypes, while the
reference panel is simulated from CEU haplotypes under the same coalescent
model. Therefore \(R_0\) is not just noisy; it is estimating the LD structure of
a different population.

The plan is:

1. Simulate one CHB GWAS sample and one CEU reference panel.
2. Build \(\bar R\) for a grid of \(\lambda\) values and fit SuSiE-relax to each.
3. Plot the final ELBO of each fit against \(\lambda\).
4. Report the PIP result of the best \(\lambda\) (largest final ELBO) and
   compare it to ordinary susieR fit on \(R_0\).

Because the ELBO is a variational lower bound on \(\log p(x \mid \bar R)\) and
\(x\) is held fixed, comparing the final ELBO across the \(\lambda\) grid is a
reasonable model-selection criterion for \(\lambda\).

```{r helper-functions}
make_rbar <- function(R0, lambda = 0.01) {
  Rbar <- (1 - lambda) * R0 + lambda * diag(1, ncol(R0))
  cov2cor(Rbar)
}

make_cs_from_alpha <- function(alpha, R, coverage = 0.95,
                               min_abs_corr = 0.5,
                               min_top_alpha = 1e-4) {
  alpha <- as.matrix(alpha)
  cs <- list()
  purity <- numeric()

  for (ell in seq_len(nrow(alpha))) {
    a <- as.numeric(alpha[ell, ])
    if (max(a) < min_top_alpha) {
      next
    }

    ord <- order(a, decreasing = TRUE)
    k <- which(cumsum(a[ord]) >= coverage)[1]
    vars <- ord[seq_len(k)]
    p <- if (length(vars) == 1) {
      1
    } else {
      R_sub <- abs(R[vars, vars, drop = FALSE])
      min(R_sub[upper.tri(R_sub)])
    }

    if (is.finite(p) && p >= min_abs_corr) {
      nm <- paste0("L", ell)
      cs[[nm]] <- vars
      purity[nm] <- p
    }
  }

  list(cs = cs, purity = purity)
}

count_selected_cs <- function(alpha, R, coverage = 0.95,
                              min_abs_corr = 0.5,
                              min_top_alpha = 1e-4) {
  length(make_cs_from_alpha(
    alpha = alpha,
    R = R,
    coverage = coverage,
    min_abs_corr = min_abs_corr,
    min_top_alpha = min_top_alpha
  )$cs)
}

last_elbo <- function(fit) {
  tail(stats::na.omit(fit$elbo), 1)
}

summarize_fit <- function(label, alpha, pip, R_for_cs, beta_true,
                          nu0 = NA_real_, sigma2 = NA_real_) {
  alpha <- as.matrix(alpha)
  top <- apply(alpha, 1, which.max)
  top_alpha <- apply(alpha, 1, max)
  cs_info <- make_cs_from_alpha(alpha, R = R_for_cs)

  data.frame(
    method = label,
    selected_L = length(cs_info$cs),
    nu0 = nu0,
    sigma2 = paste(sprintf("%.4g", sigma2), collapse = ", "),
    gamma_hat = paste(top, collapse = ", "),
    top_alpha = paste(sprintf("%.3f", top_alpha), collapse = ", "),
    causal_pip = paste(sprintf("%.3f", pip[beta_true != 0]), collapse = ", ")
  )
}

plot_pip <- function(fit, R_for_cs, beta_true, main) {
  susie_plot_alpha(
    fit,
    R = R_for_cs,
    b = beta_true,
    main = main,
    coverage = 0.95,
    min_abs_corr = 0.5
  )
  grid()
}
```

# Simulate the different-ancestry data

We use the same CHB/CEU ancestry-mismatch setup as the final section of the
SuSiE-relax demonstration. The GWAS sample comes from CHB, the reference panel
comes from CEU, and two causal SNPs are simulated in the GWAS sample.

```{r ancestry-mismatch-simulation-function}
sim_asn_eur <- function(chrom, start, end, n_gwas, n_ref, J, h2, seed) {

  stdpopsim <- reticulate::import("stdpopsim")
  msprime   <- reticulate::import("msprime")

  set.seed(seed)

  species <- stdpopsim$get_species("HomSap")
  contig  <- species$get_contig(chrom, left = as.integer(start), right = as.integer(end))

  demo_model <- species$get_demographic_model("OutOfAfrica_3G09")
  demography <- demo_model$model

  ts <- msprime$sim_ancestry(
    samples = reticulate::dict(CHB = as.integer(n_gwas), CEU = as.integer(n_ref)),
    demography = demography,
    recombination_rate = contig$recombination_map,
    random_seed = as.integer(seed)
  )

  ts <- msprime$sim_mutations(ts, rate = 1.29e-8, random_seed = as.integer(seed))

  chb_id <- NULL
  ceu_id <- NULL
  for (p in reticulate::iterate(ts$populations())) {
    if (!is.null(p$metadata) && p$metadata$name == "CHB") chb_id <- p$id
    if (!is.null(p$metadata) && p$metadata$name == "CEU") ceu_id <- p$id
  }

  samples_gwas <- ts$samples(population = as.integer(chb_id))
  samples_ref  <- ts$samples(population = as.integer(ceu_id))

  G_all <- ts$genotype_matrix()

  to_diploid <- function(hap_idx) {
    idx <- as.integer(hap_idx) + 1
    g_hap <- G_all[, idx, drop = FALSE]
    g_dip <- g_hap[, seq(1, ncol(g_hap), by = 2)] +
      g_hap[, seq(2, ncol(g_hap), by = 2)]
    t(g_dip)
  }

  G_gwas <- to_diploid(samples_gwas)
  G_ref  <- to_diploid(samples_ref)

  maf_gwas <- colMeans(G_gwas) / 2
  maf_ref  <- colMeans(G_ref) / 2

  common_mask <- (maf_gwas > 0.05) & (maf_gwas < 0.95) &
    (maf_ref > 0.05) & (maf_ref < 0.95)
  n_common <- sum(common_mask)
  if (n_common < J) {
    stop(paste0("Only ", n_common, " common variants found but J = ", J,
                ". Try a larger region."))
  }
  G_gwas <- G_gwas[, common_mask, drop = FALSE]
  G_ref  <- G_ref[, common_mask, drop = FALSE]

  J_all <- min(ncol(G_gwas), J)
  max_start <- J_all - J
  start_idx <- if (max_start > 0) sample(0:max_start, 1) else 0
  cols_to_keep <- (start_idx + 1):(start_idx + J)
  G  <- G_gwas[, cols_to_keep, drop = FALSE]
  G0 <- G_ref[, cols_to_keep, drop = FALSE]

  X <- sweep(G, 2, colMeans(G), "-")
  N <- nrow(X)

  G0_centered <- sweep(G0, 2, colMeans(G0), "-")

  R0 <- stats::cov(G0_centered)
  R0[is.na(R0)] <- 0

  R <- stats::cov(X)
  R[is.na(R)] <- 0

  num_causal <- 2
  first_causal_SNP <- sample(1:J, 1)
  second_causal_SNP <- which.min(R[first_causal_SNP, ]^2)
  causal_SNPs_loc <- c(first_causal_SNP, second_causal_SNP)

  beta_true <- rep(0, J)
  beta_true[causal_SNPs_loc] <- stats::runif(num_causal) + 0.01

  var_expl <- sum((X %*% beta_true)^2 / N)
  scale <- h2 / var_expl
  beta_true <- beta_true * sqrt(scale)

  y <- (X %*% beta_true) + sqrt(1 - h2) * stats::rnorm(N)
  y_std <- sqrt(sum((y - mean(y))^2) / N)
  y <- (y - mean(y)) / y_std

  v <- (t(X) %*% y) / N

  list(
    v = as.vector(v),
    R = R,
    R0 = R0,
    beta_true = beta_true,
    causal_SNPs_loc = causal_SNPs_loc
  )
}
```

```{r ancestry-mismatch-simulate}
seed <- 7
chrom <- "chr22"
start <- 20e6
end <- 21e6
N <- 25000L
N0 <- 1000L
J <- 500L
L <- 5L
h2 <- 0.01

res <- sim_asn_eur(
  chrom, start, end,
  n_gwas = N,
  n_ref = N0,
  J = J,
  h2 = h2,
  seed = seed
)

v <- res$v
R <- res$R
R0 <- res$R0
beta_true <- res$beta_true
causal_SNPs_loc <- res$causal_SNPs_loc

d <- 1 / sqrt(diag(R))
v <- d * v
R <- cov2cor(R)
R0 <- cov2cor(R0)

plot(v, pch = 16, xlab = "SNP index", ylab = "Marginal effect")
points(causal_SNPs_loc, v[causal_SNPs_loc], pch = 16, col = "red")

causal_SNPs_loc
```

For reference, we also fit ordinary susieR directly on the out-of-sample LD
\(R_0\). This fit does not depend on \(\lambda\).

```{r fit-susieR}
fit_susie_out <- susie_suff_stat(
  XtX = N * R0,
  Xty = N * v,
  n = N,
  yty = N,
  L = L
)
```

# Fit SuSiE-relax over a grid of lambda

```{r lambda-grid-fit}
lambda_grid <- c(1e-4, 1e-3, 1e-2, 0.05, 0.1, 0.15, 0.2, 0.3, 0.4, 0.5)

relax_fits <- lapply(lambda_grid, function(lambda) {
  Rbar <- make_rbar(R0, lambda = lambda)
  fit_susie_relax(
    x = v,
    Rbar = Rbar,
    N = N,
    L = L,
    sigma2 = rep(0.3^2, L),
    nu0_init = 2000,
    estimate_nu0 = TRUE,
    estimate_sigma2 = TRUE,
    warmup_iter = 5,
    max_iter = 3000,
    tol = 1e-7,
    verbose = FALSE,
    nu0_bounds = c(10, 10000),
    sigma2_bounds = c(1e-12, 1),
    elbo_update_interval = 5,
    nu0_update_interval = 1,
    sigma2_update_interval = 1,
    nu0_tol = 1e-4
  )
})
names(relax_fits) <- sprintf("lambda=%g", lambda_grid)

lambda_summary <- do.call(rbind, lapply(seq_along(lambda_grid), function(i) {
  fit <- relax_fits[[i]]
  data.frame(
    lambda      = lambda_grid[i],
    final_elbo  = last_elbo(fit),
    n_iter      = length(fit$elbo),
    nu0         = fit$nu0,
    tau2        = fit$tau2,
    sigma2      = paste(sprintf("%.4g", fit$sigma2), collapse = ", "),
    selected_L  = count_selected_cs(fit$alpha, R = make_rbar(R0, lambda_grid[i])),
    causal_pip  = paste(sprintf("%.3f", fit$pip[beta_true != 0]), collapse = ", ")
  )
}))

lambda_summary
```

# Final ELBO against lambda

```{r elbo-vs-lambda}
best_idx    <- which.max(lambda_summary$final_elbo)
best_lambda <- lambda_summary$lambda[best_idx]

plot(
  lambda_summary$lambda, lambda_summary$final_elbo,
  type = "b", lwd = 2, pch = 16, log = "x",
  xlab = expression(lambda), ylab = "Final ELBO",
  main = "Different ancestry: final ELBO vs lambda"
)
points(
  best_lambda, lambda_summary$final_elbo[best_idx],
  col = "red", pch = 1, cex = 2.5, lwd = 2
)
grid()

best_lambda
```

The red circle marks the \(\lambda\) with the largest final ELBO.

We also zoom in on the moderate-to-large \(\lambda\) region that contains the
value used in the demonstration.

```{r elbo-vs-lambda-zoom}
zoom_lambdas <- lambda_grid[lambda_grid >= 0.05]
zoom_summary <- lambda_summary[lambda_summary$lambda %in% zoom_lambdas, ]
zoom_best_idx <- which.max(zoom_summary$final_elbo)

plot(
  zoom_summary$lambda, zoom_summary$final_elbo,
  type = "b", lwd = 2, pch = 16,
  xlab = expression(lambda), ylab = "Final ELBO",
  main = "Different ancestry: final ELBO vs lambda (zoomed)"
)
points(
  zoom_summary$lambda[zoom_best_idx], zoom_summary$final_elbo[zoom_best_idx],
  col = "red", pch = 1, cex = 2.5, lwd = 2
)
grid()
```

# PIP of the best lambda vs susieR on R0

```{r best-summary}
best_fit  <- relax_fits[[best_idx]]
best_Rbar <- make_rbar(R0, lambda = best_lambda)

summary <- rbind(
  summarize_fit(
    label = "susieR with R0",
    alpha = fit_susie_out$alpha,
    pip = fit_susie_out$pip,
    R_for_cs = R0,
    beta_true = beta_true
  ),
  summarize_fit(
    label = sprintf("SuSiE-relax (best lambda = %g)", best_lambda),
    alpha = best_fit$alpha,
    pip = best_fit$pip,
    R_for_cs = best_Rbar,
    beta_true = beta_true,
    nu0 = best_fit$nu0,
    sigma2 = best_fit$sigma2
  )
)

summary
```

```{r best-pip-plots}
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
plot_pip(
  fit_susie_out,
  R0,
  beta_true,
  main = "Different ancestry: susieR with R0"
)
plot_pip(
  best_fit,
  best_Rbar,
  beta_true,
  main = sprintf("SuSiE-relax (lambda = %g)", best_lambda)
)
par(old_par)
```
