---
title: "PIP Comparison: SuSiE-relax vs Collapsed-R"
author: "Dat Do"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)

library(susieR)

source("analysis/compare-collapse-vs-relax-diff-ancestry.R")
source("code/plot.R")
```

# Goal

This report compares the posterior inclusion probabilities from SuSiE-relax and
the collapsed-\(R\) shared-\(q(\omega)\) method for one different-ancestry
simulation seed at a time. The main diagnostic question is whether
collapsed-\(R\) is losing power because its PIP is too diffuse.

Change `seed_to_plot` in the settings chunk below, run the single-seed fitting
chunk, and re-render the report.

```{r settings}
output_prefix <- "analysis/compare-pip-relax-vs-collapse-diff-ancestry"

# Change this one number to inspect another simulated dataset.
seed_to_plot <- 1

# The report runs only this seed if its cached result is missing. Set
# force_refit_selected_seed <- TRUE to recompute the same seed after changing
# fitting settings.
fit_selected_seed_if_missing <- TRUE
force_refit_selected_seed <- FALSE

selected_seed_prefix <- sprintf("%s-seed-%s", output_prefix, seed_to_plot)
selected_seed_rds <- paste0(selected_seed_prefix, "-diagnostics.rds")

lambda_grid <- c(1e-4, 1e-3, 1e-2, 0.05, 0.1, 0.15, 0.2, 0.3, 0.4, 0.5)
collapse_nu0_grid <- c(10, 25, 50, 100, 200, 500, 1000, 2000, 5000)
```

# Diagnostic Helpers

```{r diagnostic-helpers}
pip_entropy <- function(pip) {
  s <- sum(pip)
  if (!is.finite(s) || s <= 0) {
    return(NA_real_)
  }
  pp <- pip / s
  pp <- pp[pp > 0]
  -sum(pp * log(pp))
}

summarize_pip_fit <- function(seed, method, fit, beta_true, R_for_cs,
                              lambda, fit_seconds) {
  pip <- as.numeric(fit$pip)
  causal <- which(beta_true != 0)
  noncausal <- setdiff(seq_along(pip), causal)
  pip_rank <- rank(-pip, ties.method = "min")
  pip_norm <- pip / sum(pip)
  entropy <- pip_entropy(pip)

  data.frame(
    seed = seed,
    method = method,
    lambda = lambda,
    nu0 = fit$nu0,
    selected_L = count_selected_cs(fit$alpha, R = R_for_cs),
    total_pip = sum(pip),
    causal_pip_sum = sum(pip[causal]),
    causal_pip_min = min(pip[causal]),
    causal_pip_max = max(pip[causal]),
    causal_rank_best = min(pip_rank[causal]),
    causal_rank_worst = max(pip_rank[causal]),
    false_pip_sum = sum(pip[noncausal]),
    max_false_pip = max(pip[noncausal]),
    top1_pip = max(pip),
    top5_pip_sum = sum(sort(pip, decreasing = TRUE)[seq_len(min(5, length(pip)))]),
    top10_pip_sum = sum(sort(pip, decreasing = TRUE)[seq_len(min(10, length(pip)))]),
    n_pip_gt_001 = sum(pip > 0.01),
    n_pip_gt_005 = sum(pip > 0.05),
    n_pip_gt_010 = sum(pip > 0.10),
    n_pip_gt_050 = sum(pip > 0.50),
    effective_snps_entropy = exp(entropy),
    effective_snps_hhi = 1 / sum(pip_norm^2),
    sigma2 = paste(sprintf("%.4g", fit$sigma2), collapse = ", "),
    fit_seconds = fit_seconds
  )
}

make_pip_rows <- function(seed, method, fit, beta_true) {
  pip <- as.numeric(fit$pip)
  data.frame(
    seed = seed,
    method = method,
    snp = seq_along(pip),
    pip = pip,
    is_causal = seq_along(pip) %in% which(beta_true != 0)
  )
}

make_top_pip_rows <- function(seed, method, fit, beta_true, n_top = 20) {
  pip <- as.numeric(fit$pip)
  ord <- order(pip, decreasing = TRUE)
  ord <- ord[seq_len(min(n_top, length(ord)))]
  data.frame(
    seed = seed,
    method = method,
    rank = seq_along(ord),
    snp = ord,
    pip = pip[ord],
    is_causal = ord %in% which(beta_true != 0)
  )
}

fit_one_seed_for_pip <- function(seed, chrom = "chr22", start = 20e6,
                                 end = 21e6, N = 25000L, N0 = 500L,
                                 J = 500L, h2 = 0.01, L = 5,
                                 lambda_grid = c(1e-4, 1e-3, 1e-2, 0.05,
                                                 0.1, 0.15, 0.2, 0.3, 0.4,
                                                 0.5),
                                 collapse_nu0_grid = c(10, 25, 50, 100, 200,
                                                       500, 1000, 2000, 5000),
                                 relax_max_iter = 3000,
                                 collapse_max_iter = 200,
                                 verbose = TRUE) {
  if (verbose) {
    message(sprintf("seed %d: simulate", seed))
  }
  sim_time <- system.time({
    res <- sim_asn_eur(
      chrom = chrom,
      start = start,
      end = end,
      n_gwas = N,
      n_ref = N0,
      J = J,
      h2 = h2,
      seed = seed
    )
    res <- standardize_diff_ancestry_sim(res)
  })

  x <- res$v
  R0 <- res$R0
  beta_true <- res$beta_true

  if (verbose) {
    message(sprintf("seed %d: fit SuSiE-relax lambda grid", seed))
  }
  relax <- fit_relax_lambda_grid(
    x = x,
    R0 = R0,
    N = N,
    L = L,
    beta_true = beta_true,
    lambda_grid = lambda_grid,
    max_iter = relax_max_iter
  )

  if (verbose) {
    message(sprintf("seed %d: fit collapsed-R lambda grid", seed))
  }
  collapse <- fit_collapse_lambda_grid(
    x = x,
    R0 = R0,
    N = N,
    L = L,
    beta_true = beta_true,
    lambda_grid = lambda_grid,
    nu0_grid = collapse_nu0_grid,
    max_iter = collapse_max_iter
  )

  relax_rbar <- make_rbar(R0, relax$best_lambda)
  collapse_rbar <- make_rbar(R0, collapse$best_lambda)

  metrics <- rbind(
    summarize_pip_fit(
      seed = seed,
      method = "relax",
      fit = relax$best_fit,
      beta_true = beta_true,
      R_for_cs = relax_rbar,
      lambda = relax$best_lambda,
      fit_seconds = relax$fit_seconds
    ),
    summarize_pip_fit(
      seed = seed,
      method = "collapse_shared",
      fit = collapse$best_fit,
      beta_true = beta_true,
      R_for_cs = collapse_rbar,
      lambda = collapse$best_lambda,
      fit_seconds = collapse$fit_seconds
    )
  )
  metrics$causal <- paste(which(beta_true != 0), collapse = ", ")
  metrics$sim_seconds <- unname(sim_time["elapsed"])

  pips <- rbind(
    make_pip_rows(seed, "relax", relax$best_fit, beta_true),
    make_pip_rows(seed, "collapse_shared", collapse$best_fit, beta_true)
  )

  top_pips <- rbind(
    make_top_pip_rows(seed, "relax", relax$best_fit, beta_true),
    make_top_pip_rows(seed, "collapse_shared", collapse$best_fit, beta_true)
  )

  best <- rbind(
    transform(relax$profile[relax$best_idx, ], seed = seed),
    transform(collapse$profile[collapse$best_idx, ], seed = seed)
  )

  list(
    seed = seed,
    metrics = metrics,
    pips = pips,
    top_pips = top_pips,
    best = best,
    profiles = rbind(
      transform(relax$profile, seed = seed),
      transform(collapse$profile, seed = seed)
    )
  )
}

combine_pip_runs <- function(runs) {
  list(
    metrics = do.call(rbind, lapply(runs, `[[`, "metrics")),
    pips = do.call(rbind, lapply(runs, `[[`, "pips")),
    top_pips = do.call(rbind, lapply(runs, `[[`, "top_pips")),
    best = do.call(rbind, lapply(runs, `[[`, "best")),
    profiles = do.call(rbind, lapply(runs, `[[`, "profiles")),
    runs = runs
  )
}

run_pip_diagnostics <- function(seeds = seed_to_plot,
                                output_prefix =
                                  "analysis/compare-pip-relax-vs-collapse-diff-ancestry",
                                ...) {
  dir.create(dirname(output_prefix), recursive = TRUE, showWarnings = FALSE)

  runs <- vector("list", length(seeds))
  names(runs) <- paste0("seed", seeds)

  for (i in seq_along(seeds)) {
    runs[[i]] <- fit_one_seed_for_pip(seed = seeds[i], ...)
    partial <- combine_pip_runs(runs[seq_len(i)])
    saveRDS(partial, paste0(output_prefix, "-diagnostics.rds"))
    utils::write.csv(
      partial$metrics,
      paste0(output_prefix, "-metrics.csv"),
      row.names = FALSE
    )
    utils::write.csv(
      partial$top_pips,
      paste0(output_prefix, "-top-pips.csv"),
      row.names = FALSE
    )
    utils::write.csv(
      partial$pips,
      paste0(output_prefix, "-pips.csv"),
      row.names = FALSE
    )
  }

  combine_pip_runs(runs)
}

load_pip_diagnostics <- function(path = selected_seed_rds) {
  if (file.exists(path)) {
    readRDS(path)
  } else {
    list(
      metrics = data.frame(),
      pips = data.frame(),
      top_pips = data.frame(),
      best = data.frame(),
      profiles = data.frame(),
      runs = list()
    )
  }
}

plot_missing_diagnostics <- function() {
  plot.new()
  text(
    0.5, 0.5,
    paste(
      sprintf("No cached diagnostics found for seed %s.", seed_to_plot),
      "Run the single-seed fitting chunk first.",
      sep = "\n"
    )
  )
}
```

# Fit Or Load Selected Seed

This chunk fits only `seed_to_plot` if the cached RDS for that seed is missing.
To inspect another dataset, change `seed_to_plot` in the settings chunk and
re-knit.

```{r fit-or-load-selected-seed}
if (force_refit_selected_seed ||
    (!file.exists(selected_seed_rds) && fit_selected_seed_if_missing)) {
  pip_diag <- run_pip_diagnostics(
    seeds = seed_to_plot,
    output_prefix = selected_seed_prefix,
    lambda_grid = lambda_grid,
    collapse_nu0_grid = collapse_nu0_grid
  )
} else {
  pip_diag <- load_pip_diagnostics()
}

have_diagnostics <- nrow(pip_diag$metrics) > 0
have_diagnostics
```

For a faster exploratory view, set smaller iteration limits in the call above,
for example `relax_max_iter = 500` and `collapse_max_iter = 50`.

# Selected Seed Summary

```{r selected-seed-summary}
if (have_diagnostics) {
  pip_diag$metrics[
    ,
    c(
      "seed", "method", "lambda", "nu0", "selected_L", "total_pip",
      "causal_pip_sum", "false_pip_sum", "max_false_pip",
      "effective_snps_entropy", "effective_snps_hhi", "n_pip_gt_001",
      "n_pip_gt_005", "n_pip_gt_010", "causal"
    )
  ]
} else {
  cat("No cached diagnostics yet.\n")
}
```

```{r lambda-profile}
if (have_diagnostics) {
  pip_diag$profiles[
    order(pip_diag$profiles$method, pip_diag$profiles$lambda),
    c("seed", "method", "lambda", "final_elbo", "n_iter", "nu0",
      "selected_L", "causal_pip", "top", "top_alpha")
  ]
} else {
  cat("No cached diagnostics yet.\n")
}
```

# PIP Plots

The red points and rug marks are the two true causal SNPs.

```{r selected-seed-pip-plots}
if (have_diagnostics) {
  pip_seed <- pip_diag$pips[pip_diag$pips$seed == seed_to_plot, ]
  y_max <- max(pip_seed$pip)

  old_par <- par(no.readonly = TRUE)
  par(mfrow = c(2, 1), mar = c(4, 4, 3, 1))

  for (method_i in c("relax", "collapse_shared")) {
    d <- pip_seed[pip_seed$method == method_i, ]
    metric_i <- pip_diag$metrics[pip_diag$metrics$method == method_i, ]
    plot(
      d$snp, d$pip,
      pch = 16,
      cex = 0.7,
      ylim = c(0, y_max),
      xlab = "SNP",
      ylab = "PIP",
      main = sprintf(
        "Seed %d: %s, lambda=%g, nu0=%.3g, selected L=%d",
        seed_to_plot,
        method_i,
        metric_i$lambda,
        metric_i$nu0,
        metric_i$selected_L
      )
    )
    points(d$snp[d$is_causal], d$pip[d$is_causal], col = 2, pch = 16)
    rug(d$snp[d$is_causal], col = 2, lwd = 1.5)
    grid()
  }

  par(old_par)
} else {
  plot_missing_diagnostics()
}
```

# Diffusion View

If collapsed-\(R\) is too diffuse, it should have more small nonzero PIPs and a
larger effective number of SNPs.

```{r sorted-pip-plots}
if (have_diagnostics) {
  pip_seed <- pip_diag$pips[pip_diag$pips$seed == seed_to_plot, ]

  old_par <- par(no.readonly = TRUE)
  par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

  for (method_i in c("relax", "collapse_shared")) {
    d <- pip_seed[pip_seed$method == method_i, ]
    sorted_pip <- sort(d$pip, decreasing = TRUE)
    plot(
      seq_along(sorted_pip),
      sorted_pip,
      type = "l",
      lwd = 2,
      log = "y",
      xlab = "PIP rank",
      ylab = "PIP",
      main = sprintf("Seed %d: sorted PIP, %s", seed_to_plot, method_i)
    )
    grid()
  }

  par(old_par)
} else {
  plot_missing_diagnostics()
}
```

```{r top-pip-table}
if (have_diagnostics) {
  pip_diag$top_pips
} else {
  cat("No cached diagnostics yet.\n")
}
```

# PIP Threshold Counts

```{r pip-threshold-barplot}
if (have_diagnostics) {
  old_par <- par(no.readonly = TRUE)

  threshold_counts <- pip_diag$metrics[
    ,
    c("method", "n_pip_gt_001", "n_pip_gt_005", "n_pip_gt_010", "n_pip_gt_050")
  ]
  rownames(threshold_counts) <- threshold_counts$method
  count_matrix <- t(as.matrix(threshold_counts[, -1]))

  barplot(
    count_matrix,
    beside = TRUE,
    legend.text = rownames(count_matrix),
    args.legend = list(bty = "n", x = "topright"),
    ylab = "# SNPs",
    main = sprintf("Seed %d: number of SNPs above PIP thresholds", seed_to_plot)
  )
  grid()
  threshold_counts

  par(old_par)
} else {
  plot_missing_diagnostics()
}
```
