---
title: "Scale of q as lambda varies"
author: "Dat Do"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)

source("code/sim_2pop.R")
```

# Goal

In the SER-relax calculation, the term
\[
q(\lambda) = x^\top \bar R(\lambda)^{-1} x
\]
can be important, where
\[
\bar R(\lambda) = (1-\lambda)R_0 + \lambda I.
\]
This note investigates the scale of \(q(\lambda)\) in three settings:

1. Low mismatch: same ancestry, large reference panel.
2. High mismatch from small reference sample size.
3. High mismatch from different ancestry.

The goal is only to understand how \(q\) changes with \(\lambda\), not to fit
SuSiE-relax.

```{r helpers}
standardize_sim <- function(res) {
  d <- 1 / sqrt(diag(res$R))
  res$v <- d * res$v
  res$R <- cov2cor(res$R)
  res$R0 <- cov2cor(res$R0)
  res
}

q_path <- function(x, R0, lambda_grid) {
  R0 <- (R0 + t(R0)) / 2
  eig <- eigen(R0, symmetric = TRUE)
  x_rot <- crossprod(eig$vectors, x)

  out <- do.call(rbind, lapply(lambda_grid, function(lambda) {
    denom <- (1 - lambda) * eig$values + lambda
    data.frame(
      lambda = lambda,
      q = sum(as.numeric(x_rot)^2 / denom),
      min_eigen = min(denom),
      max_eigen = max(denom),
      condition_number = max(denom) / min(denom),
      n_nonpositive_eigen = sum(denom <= 0)
    )
  }))
  out
}

q_from_R <- function(x, R) {
  R <- (R + t(R)) / 2
  as.numeric(crossprod(x, solve(R, x)))
}

plot_q_path <- function(tab, main, q_true = NULL) {
  plot(
    tab$lambda, tab$q,
    type = "l", lwd = 2, log = "x",
    xlab = expression(lambda),
    ylab = expression(q(lambda) == x^T * bar(R)(lambda)^{-1} * x),
    main = main
  )
  points(tab$lambda, tab$q, pch = 16, cex = 0.6)
  if (!is.null(q_true)) {
    abline(h = q_true, col = "gray40", lty = 2, lwd = 2)
  }
  grid()
}

plot_condition_path <- function(tab, main) {
  plot(
    tab$lambda, tab$condition_number,
    type = "l", lwd = 2, log = "xy",
    xlab = expression(lambda),
    ylab = "Condition number",
    main = main
  )
  points(tab$lambda, tab$condition_number, pch = 16, cex = 0.6)
  grid()
}

sim_asn_eur <- function(chrom, start, end, n_gwas, n_ref, J, h2, seed) {

  stdpopsim <- reticulate::import("stdpopsim")
  msprime   <- reticulate::import("msprime")

  set.seed(seed)

  species <- stdpopsim$get_species("HomSap")
  contig  <- species$get_contig(chrom, left = as.integer(start), right = as.integer(end))

  demo_model <- species$get_demographic_model("OutOfAfrica_3G09")
  demography <- demo_model$model

  ts <- msprime$sim_ancestry(
    samples = reticulate::dict(CHB = as.integer(n_gwas), CEU = as.integer(n_ref)),
    demography = demography,
    recombination_rate = contig$recombination_map,
    random_seed = as.integer(seed)
  )

  ts <- msprime$sim_mutations(ts, rate = 1.29e-8, random_seed = as.integer(seed))

  chb_id <- NULL
  ceu_id <- NULL
  for (p in reticulate::iterate(ts$populations())) {
    if (!is.null(p$metadata) && p$metadata$name == "CHB") chb_id <- p$id
    if (!is.null(p$metadata) && p$metadata$name == "CEU") ceu_id <- p$id
  }

  samples_gwas <- ts$samples(population = as.integer(chb_id))
  samples_ref  <- ts$samples(population = as.integer(ceu_id))

  G_all <- ts$genotype_matrix()

  to_diploid <- function(hap_idx) {
    idx <- as.integer(hap_idx) + 1
    g_hap <- G_all[, idx, drop = FALSE]
    g_dip <- g_hap[, seq(1, ncol(g_hap), by = 2)] +
      g_hap[, seq(2, ncol(g_hap), by = 2)]
    t(g_dip)
  }

  G_gwas <- to_diploid(samples_gwas)
  G_ref  <- to_diploid(samples_ref)

  maf_gwas <- colMeans(G_gwas) / 2
  maf_ref  <- colMeans(G_ref) / 2

  common_mask <- (maf_gwas > 0.05) & (maf_gwas < 0.95) &
    (maf_ref > 0.05) & (maf_ref < 0.95)
  n_common <- sum(common_mask)
  if (n_common < J) {
    stop(paste0("Only ", n_common, " common variants found but J = ", J,
                ". Try a larger region."))
  }
  G_gwas <- G_gwas[, common_mask, drop = FALSE]
  G_ref  <- G_ref[, common_mask, drop = FALSE]

  J_all <- min(ncol(G_gwas), J)
  max_start <- J_all - J
  start_idx <- if (max_start > 0) sample(0:max_start, 1) else 0
  cols_to_keep <- (start_idx + 1):(start_idx + J)
  G  <- G_gwas[, cols_to_keep, drop = FALSE]
  G0 <- G_ref[, cols_to_keep, drop = FALSE]

  X <- sweep(G, 2, colMeans(G), "-")
  N <- nrow(X)

  G0_centered <- sweep(G0, 2, colMeans(G0), "-")

  R0 <- stats::cov(G0_centered)
  R0[is.na(R0)] <- 0

  R <- stats::cov(X)
  R[is.na(R)] <- 0

  num_causal <- 2
  first_causal_SNP <- sample(1:J, 1)
  second_causal_SNP <- which.min(R[first_causal_SNP, ]^2)
  causal_SNPs_loc <- c(first_causal_SNP, second_causal_SNP)

  beta_true <- rep(0, J)
  beta_true[causal_SNPs_loc] <- stats::runif(num_causal) + 0.01

  var_expl <- sum((X %*% beta_true)^2 / N)
  scale <- h2 / var_expl
  beta_true <- beta_true * sqrt(scale)

  y <- (X %*% beta_true) + sqrt(1 - h2) * stats::rnorm(N)
  y_std <- sqrt(sum((y - mean(y))^2) / N)
  y <- (y - mean(y)) / y_std

  v <- (t(X) %*% y) / N

  list(
    v = as.vector(v),
    R = R,
    R0 = R0,
    beta_true = beta_true,
    causal_SNPs_loc = causal_SNPs_loc
  )
}
```

```{r lambda-grid}
lambda_grid <- c(10^seq(-6, -1, length.out = 45), seq(0.12, 1, length.out = 23))
```

# Simulate the three settings

```{r simulate-low-mismatch}
res_low <- sim_2pop(
  chrom = "chr22",
  start = 20e6,
  end = 21e6,
  T_split = 20,
  n_gwas = 20000L,
  n_ref = 2000L,
  J = 1000L,
  h2 = 0.01,
  seed = 6
)
res_low <- standardize_sim(res_low)
```

```{r simulate-small-reference-mismatch}
res_small_ref <- sim_2pop(
  chrom = "chr22",
  start = 20e6,
  end = 21e6,
  T_split = 20,
  n_gwas = 20000L,
  n_ref = 100L,
  J = 1000L,
  h2 = 0.01,
  seed = 9
)
res_small_ref <- standardize_sim(res_small_ref)
```

```{r simulate-ancestry-mismatch}
res_ancestry <- sim_asn_eur(
  chrom = "chr22",
  start = 20e6,
  end = 21e6,
  n_gwas = 25000L,
  n_ref = 500L,
  J = 500L,
  h2 = 0.01,
  seed = 2
)
res_ancestry <- standardize_sim(res_ancestry)
```

# Compute q over lambda

```{r compute-q}
q_low <- q_path(res_low$v, res_low$R0, lambda_grid)
q_low$setting <- "Low mismatch"

q_small_ref <- q_path(res_small_ref$v, res_small_ref$R0, lambda_grid)
q_small_ref$setting <- "Small reference mismatch"

q_ancestry <- q_path(res_ancestry$v, res_ancestry$R0, lambda_grid)
q_ancestry$setting <- "Ancestry mismatch"

q_all <- rbind(q_low, q_small_ref, q_ancestry)

q_true <- c(
  "Low mismatch" = q_from_R(res_low$v, res_low$R),
  "Small reference mismatch" = q_from_R(res_small_ref$v, res_small_ref$R),
  "Ancestry mismatch" = q_from_R(res_ancestry$v, res_ancestry$R)
)

summary_table <- do.call(rbind, lapply(split(q_all, q_all$setting), function(tab) {
  data.frame(
    setting = tab$setting[1],
    q_in_sample_R = q_true[tab$setting[1]],
    q_min = min(tab$q),
    q_max = max(tab$q),
    q_lambda_1e_4 = tab$q[which.min(abs(tab$lambda - 1e-4))],
    q_lambda_1e_2 = tab$q[which.min(abs(tab$lambda - 1e-2))],
    q_lambda_0_1 = tab$q[which.min(abs(tab$lambda - 0.1))],
    q_lambda_1 = tab$q[which.min(abs(tab$lambda - 1))]
  )
}))

summary_table
```

# q as lambda varies

```{r q-plots, fig.width=10, fig.height=4}
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 3))
plot_q_path(q_low, "Low mismatch", q_true["Low mismatch"])
plot_q_path(q_small_ref, "Small reference mismatch", q_true["Small reference mismatch"])
plot_q_path(q_ancestry, "Ancestry mismatch", q_true["Ancestry mismatch"])
par(old_par)
```

The dashed horizontal line is \(x^\top R^{-1}x\), using the in-sample LD
matrix \(R\). This is not available in real summary-data analysis, but it is a
useful simulation reference.

```{r q-combined, fig.width=7, fig.height=5}
plot(
  range(q_all$lambda), range(q_all$q),
  type = "n", log = "xy",
  xlab = expression(lambda),
  ylab = expression(q(lambda) == x^T * bar(R)(lambda)^{-1} * x),
  main = expression("Scale of " * q(lambda) * " across mismatch settings")
)
cols <- c(
  "Low mismatch" = "black",
  "Small reference mismatch" = "red3",
  "Ancestry mismatch" = "blue3"
)
for (setting in names(cols)) {
  tab <- q_all[q_all$setting == setting, ]
  lines(tab$lambda, tab$q, lwd = 2, col = cols[setting])
  abline(h = q_true[setting], col = cols[setting], lwd = 1, lty = 2)
}
legend("topright", legend = names(cols), col = cols, lwd = 2, bty = "n")
grid()
```

# Conditioning of the stabilized LD

When \(\lambda\) is very small, \(\bar R(\lambda)\) can be poorly conditioned,
especially when the reference panel is small and \(R_0\) is rank deficient.
This explains why \(q(\lambda)\) can become very large near \(\lambda=0\).

```{r condition-plots, fig.width=10, fig.height=4}
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 3))
plot_condition_path(q_low, "Low mismatch")
plot_condition_path(q_small_ref, "Small reference mismatch")
plot_condition_path(q_ancestry, "Ancestry mismatch")
par(old_par)
```

```{r eigen-summary}
eigen_summary <- do.call(rbind, lapply(split(q_all, q_all$setting), function(tab) {
  data.frame(
    setting = tab$setting[1],
    min_eigen_at_smallest_lambda = tab$min_eigen[which.min(tab$lambda)],
    condition_at_smallest_lambda = tab$condition_number[which.min(tab$lambda)],
    min_eigen_at_lambda_0_1 = tab$min_eigen[which.min(abs(tab$lambda - 0.1))],
    condition_at_lambda_0_1 = tab$condition_number[which.min(abs(tab$lambda - 0.1))]
  )
}))

eigen_summary
```
