---
title: "SER-relax-prime PIP Diffuseness"
author: "Dat Do"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 4.5
)

source("code/sim_2pop.R")
source("code/SER-relax.R")
source("code/SuSiE-relax.R")
```

# Goal

For $L=1$, SER/SuSiE-RSS and SER-relax-prime both rank SNPs only by
$x_j^2$. The only difference is the coefficient multiplying $N x_j^2/2$.
This page uses the coalescent simulation to generate $x$ and $R_0$, sets
\[
\sigma_0^2 = \max_{j:\beta_j\ne 0}\beta_j^2,
\]
and compares the closed-form PIPs:
\[
\mathrm{PIP}_{\mathrm{rss},j}
\propto
\exp\left\{
\frac{N}{2}
\frac{N\sigma_0^2}{1+N\sigma_0^2}
x_j^2
\right\},
\]
and
\[
\mathrm{PIP}_{\mathrm{prime},j}
\propto
\exp\left\{
\frac{N}{2}
\left(
\frac{N\sigma_0^2}{1+N\sigma_0^2}
-
\frac{N\sigma_0^2\tau^2}{1+N\sigma_0^2\tau^2}
\right)
x_j^2
\right\},
\qquad
\tau^2=(\nu_0+3)^{-1}.
\]

The comparison uses $\nu_0=10$ and $\nu_0=100$. The reference LD $R_0$
is still generated by the simulation, but it does not enter these
closed-form $L=1$ PIPs because the $R_0$-dependent quadratic term is
common across candidate causal SNPs and cancels in the normalization.

```{r helper-functions}
standardize_sim <- function(res) {
  d <- 1 / sqrt(diag(res$R))
  res$v <- d * res$v
  res$R <- cov2cor(res$R)
  res$R0 <- cov2cor(res$R0)
  res
}

softmax <- function(x) {
  z <- x - max(x)
  exp_z <- exp(z)
  exp_z / sum(exp_z)
}

ser_rss_pip <- function(x, N, sigma2) {
  kappa <- N * sigma2 / (1 + N * sigma2)
  softmax(0.5 * N * kappa * x^2)
}

ser_relax_prime_pip <- function(x, N, sigma2, nu0) {
  tau2 <- 1 / (nu0 + 3)
  kappa <- N * sigma2 / (1 + N * sigma2)
  rho <- N * sigma2 * tau2 / (1 + N * sigma2 * tau2)
  softmax(0.5 * N * (kappa - rho) * x^2)
}

coefficient_table <- function(N, sigma2, nu0_grid) {
  t <- N * sigma2
  rss <- t / (1 + t)
  prime <- vapply(nu0_grid, function(nu0) {
    tau2 <- 1 / (nu0 + 3)
    rss - t * tau2 / (1 + t * tau2)
  }, numeric(1))
  data.frame(
    method = c("SER/SuSiE-RSS", paste0("SER-relax-prime, nu0=", nu0_grid)),
    nu0 = c(NA_real_, nu0_grid),
    tau2 = c(0, 1 / (nu0_grid + 3)),
    exponent_coefficient = c(rss, prime),
    relative_to_rss = c(1, prime / rss)
  )
}

pip_summary <- function(pip, label, rss_pip, causal) {
  active <- pip > 0
  entropy <- -sum(pip[active] * log(pip[active]))
  data.frame(
    method = label,
    top_snp = which.max(pip),
    top_pip = max(pip),
    causal_total_pip = sum(pip[causal]),
    effective_number = exp(entropy),
    max_abs_diff_from_rss = max(abs(pip - rss_pip)),
    corr_with_rss = suppressWarnings(cor(pip, rss_pip))
  )
}

plot_pips <- function(pip_list, causal, beta_true) {
  old_par <- par(no.readonly = TRUE)
  on.exit(par(old_par), add = TRUE)

  ymax <- max(unlist(pip_list))
  par(mfrow = c(1, length(pip_list)), mar = c(4, 4, 3, 1))
  for (nm in names(pip_list)) {
    pip <- pip_list[[nm]]
    plot(
      seq_along(pip), pip,
      type = "h",
      lwd = 1.5,
      ylim = c(0, ymax),
      xlab = "SNP index",
      ylab = "PIP",
      main = nm
    )
    points(causal, pip[causal], pch = 19, col = "red")
    rug(causal, col = "red", lwd = 2)
    grid()
  }
}
```

# Simulation

```{r simulate}
N <- 20000L
J <- 1000L
N0 <- 100L
seed <- 6
nu0_grid <- c(10, 100)

res <- sim_2pop(
  chrom = "chr22",
  start = 20e6,
  end = 21e6,
  T_split = 20,
  n_gwas = N,
  n_ref = N0,
  J = J,
  h2 = 0.01,
  seed = seed
)
res <- standardize_sim(res)

x <- res$v
R0 <- res$R0
beta_true <- res$beta_true
causal <- which(beta_true != 0)
sigma2 <- max(beta_true[causal]^2)

data.frame(
  causal_snp = causal,
  beta_true = beta_true[causal],
  beta2 = beta_true[causal]^2
)

sigma2
coefficient_table(N, sigma2, nu0_grid)
```

# Closed-form PIPs

```{r compute-pips}
pip_rss <- ser_rss_pip(x, N, sigma2)
pip_prime <- lapply(nu0_grid, function(nu0) {
  ser_relax_prime_pip(x, N, sigma2, nu0)
})
names(pip_prime) <- paste0("SER-relax-prime\nnu0=", nu0_grid)

pip_list <- c(list("SER/SuSiE-RSS" = pip_rss), pip_prime)

do.call(rbind, c(
  list(pip_summary(pip_rss, "SER/SuSiE-RSS", pip_rss, causal)),
  Map(function(pip, nm) pip_summary(pip, nm, pip_rss, causal),
      pip_prime, names(pip_prime))
))
```

```{r plot-pips}
plot_pips(pip_list, causal, beta_true)
```

# SER-relax PIPs

Unlike SER-relax-prime, original SER-relax does not have a closed-form
PIP that is simply a rescaled softmax of $x_j^2$. Here we apply the
single-effect SER-relax variational algorithm to the same simulated
data, fixing $\nu_0=10$ and $\nu_0=100$, and compare its PIPs to the
closed-form curves above.

```{r fit-ser-relax}
J = dim(R0)[1]
Rbar = .99 * R0 + .01 * diag(1, J)
fit_ser <- lapply(nu0_grid, function(nu0) {
  fit_ser_relax(
    x = x,
    Rbar = Rbar,
    N = N,
    sigma0 = sqrt(sigma2),
    nu0_init = nu0,
    estimate_nu0 = FALSE,
    max_iter = 200,
    tol = 1e-8,
    verbose = FALSE
  )
})
names(fit_ser) <- paste0("SER-relax\nnu0=", nu0_grid)

pip_ser <- lapply(fit_ser, `[[`, "pi")

do.call(rbind, Map(function(pip, nm) {
  pip_summary(pip, nm, pip_rss, causal)
}, pip_ser, names(pip_ser)))
```

```{r plot-ser-relax-pips}
plot_pips(pip_ser, causal, beta_true)
```

```{r plot-all-pips}
plot_pips(c(pip_list, pip_ser), causal, beta_true)
```

# Comment on the results

1.  We see that susie-relax-prime has near similar PIP to susie-rss when
    $\nu_0$ is around 100, while it will be too diffuse when $\nu_0$ is
    as small as 10.

2.  Looking back at the `susie-relax-prime-demonstration.html`, the
    highly mismatch case, we observe that the optimal $\nu_0$ converge
    to 600, marking that the final PIP will be similar to susie-rss's
    PIP.

3.  Looking back at the `susie-relax-demonstration.html`, the highly
    mismatch case, we observe that the optimal $\nu_0$ converge to the
    lower bound (10). But interestingly the PIP of that model is not
    diffused at all. It is still concentrated around the strongest
    associated SNP.

4.  To explain this difference between the two model: Recall the
    conditional probabilities in the two models, with the probability in
    `susie-relax` being denoted by $p$ and `susie-relax-prime` being
    denoted by $\tilde p$.

\[
p(x | \beta, \gamma = j) =
\mathcal{N}(x | \beta u_j(\mu_j), \tfrac{R}{N} + \beta^2 \tau^2 S_j^{pad})
\]
and

\[
\tilde p(x | \beta, \gamma = j) =
\mathcal{N}(x | \beta u_j(\mu_j), \tfrac{R}{N} + \sigma_0^2 \tau^2 S_j^{pad}),
\]
where

-   $\mu_j = \bar R_{j, -j}\in \mathbb{R}^{J-1}$,
-   $u_j(z_j\in \mathbb{R}^{J})$ is the vector with $j$-th coordinate
    being 1 and the rest being $\mu_j$,
-   $S_j \mathbb{R}^{(J-1)\times (J-1)}$ is the Schur complement of
    $\bar R_{jj}$
-   and $S_j^{pad} \in \mathbb{R}^{J\times J}$ is the matrix with $j$-th
    row and column being 0 and the rest being $S_j$.

Let $q = x^\top (\bar R)^{-1} x$, we have

\[
p(\gamma=j \mid x) \propto
\int
(1+N\beta^2\tau^2)^{-(J-1)/2}
\exp\!\left(-\frac{N}{2(1+N\beta^2\tau^2)}q\right) \times
\exp\!\left(-\frac{N^2\beta^2\tau^2}{2(1+N\beta^2\tau^2)}x_j^2 + N\beta x_j - \frac{N}{2}\beta^2\right)
p(\beta \mid 0,\sigma_0^2)\,d\beta.
\]
On the other hand,
\[
\tilde p(\gamma=j \mid x) \propto
\int
(1+N\sigma_0^2\tau^2)^{-(J-1)/2}
\exp\!\left(-\frac{N}{2(1+N\sigma_0^2\tau^2)}q\right) \times
\exp\!\left(-\frac{N^2\sigma_0^2\tau^2}{2(1+N\sigma_0^2\tau^2)}x_j^2 + N\beta x_j - \frac{N}{2}\beta^2\right)
p(\beta \mid 0,\sigma_0^2)\,d\beta, 
\]
Note that $\sigma_0^2$ here is common for all $j$ (and it does not
depend on $\beta$ so it can be brought out of the integral), therefore,
\[
\tilde p(\gamma=j \mid x) \propto
\int
\exp\!\left(-\frac{N^2\sigma_0^2\tau^2}{2(1+N\sigma_0^2\tau^2)}x_j^2 + N\beta x_j - \frac{N}{2}\beta^2\right)
p(\beta \mid 0,\sigma_0^2)\,d\beta, 
\]

which leads to

\[
\tilde p(\gamma=j \mid x) \propto \exp\left\{
\frac{N}{2}
\left(
\frac{N\sigma_0^2}{1+N\sigma_0^2}
-
\frac{N\sigma_0^2\tau^2}{1+N\sigma_0^2\tau^2}
\right)
x_j^2
\right\}.
\]

5.  The important distinction is not that the original `susie-relax`
    relaxes low-Z SNPs more. It is closer to the opposite. In the
    original model, the amount of relaxation is tied to the latent
    effect size through $N\beta^2\tau^2$. For a candidate SNP with large
    $x_j$, the factor
    \[
    \exp\left(N\beta x_j-\frac{N}{2}\beta^2\right)
    p(\beta\mid 0,\sigma_0^2)
    \]
    pulls the integral toward $\beta$ values with the same sign as
    $x_j$ and with magnitude roughly proportional to $|x_j|$. Thus
    large-Z SNPs are precisely the SNPs for which the integral places
    mass on larger $\beta^2$, and therefore on a larger effective
    relaxation $N\beta^2\tau^2$. Null SNPs with $x_j\approx 0$ keep most
    of their integral near $\beta\approx 0$, so their effective
    relaxation is small.

    Another way to see this is to write the original local evidence as
    \[
    p(\gamma=j\mid x)
    \propto
    \int
    H(\beta^2)
    \exp\left\{
    N\beta x_j
    -
    \frac{N}{2}\beta^2
    -
    A(\beta^2)x_j^2
    \right\}
    p(\beta\mid 0,\sigma_0^2)\,d\beta,
    \]
    where
    \[
    H(\beta^2)
    =
    (1+N\beta^2\tau^2)^{-(J-1)/2}
    \exp\left\{-\frac{N}{2(1+N\beta^2\tau^2)}q\right\}
    \]
    and
    \[
    A(\beta^2)
    =
    \frac{N^2\beta^2\tau^2}{2(1+N\beta^2\tau^2)}.
    \]
    The $H(\beta^2)$ term is common as a function of $\beta^2$, but
    it cannot be cancelled outside the integral because the linear term
    $N\beta x_j$ changes how each SNP weights different values of
    $\beta^2$. Consequently, the original model does not simply apply
    one common flattening coefficient to all $x_j^2$. The active SNP can
    get both a large effect-size posterior and the LD-column flexibility
    needed to explain the residual LD pattern. This is why a small
    $\nu_0$ can still lead to concentrated PIPs.

6.  In `susie-relax-prime`, we replace $\beta^2$ in the covariance by
    the fixed prior scale $\sigma_0^2$. Therefore the determinant term
    and the $q=x^\top\bar R^{-1}x$ term become
    \[
    (1+N\sigma_0^2\tau^2)^{-(J-1)/2}
    \exp\left\{
    -\frac{N}{2(1+N\sigma_0^2\tau^2)}q
    \right\},
    \]
    which is common across $j$ and can be pulled out of the
    normalization. What remains is exactly
    \[
    \tilde p(\gamma=j\mid x)
    \propto
    \exp\left\{
    \frac{N}{2}
    \left[
    \frac{N\sigma_0^2}{1+N\sigma_0^2}
    -
    \frac{N\sigma_0^2\tau^2}{1+N\sigma_0^2\tau^2}
    \right]
    x_j^2
    \right\}.
    \]
    Thus prime applies the same relaxation penalty to every SNP,
    independent of how strongly the data support a large $\beta$ for
    that SNP. When $\nu_0$ is small, $\tau^2$ is large and the
    subtraction term can be large, so the coefficient multiplying
    $x_j^2$ shrinks and the PIPs become more diffuse. This is a
    deliberate price paid by the prime approximation to obtain the
    irrelevance-of-null-SNPs property: for $L=1$, PIPs depend only on
    $x_j^2$, not on coordinate-specific LD-column fit.

7.  This also explains the empirical difference between the two
    demonstration pages. In the original `susie-relax` high-mismatch
    example, learning $\nu_0=10$ means that the model wants substantial
    LD-column flexibility for effects that the data support. It does not
    imply diffuse PIPs. In the `susie-relax-prime` high-mismatch
    example, learning a much larger $\nu_0$ is natural because a small
    $\nu_0$ would impose a global flattening of the $x_j^2$ evidence and
    would make the single-effect posterior too diffuse.

# SuSiE-relax With $L=2$

The single-effect SER-relax fit above can behave differently from the
previous multi-effect analysis. To check the same data under the
multi-effect model, fit SuSiE-relax with $L=2$, estimate $\nu_0$, and
plot the resulting marginal PIP.

```{r fit-susie-relax-l2}
fit_susie_relax_l2 <- fit_susie_relax(
  x = x,
  Rbar = Rbar,
  N = N,
  L = 2,
  sigma2 = rep(sigma2, 2),
  nu0_init = 2000,
  estimate_nu0 = TRUE,
  estimate_sigma2 = TRUE,
  warmup_iter = 5,
  max_iter = 1000,
  tol = 1e-7,
  verbose = FALSE,
  nu0_bounds = c(10, 10000),
  sigma2_bounds = c(1e-12, 1),
  elbo_update_interval = 10,
  nu0_update_interval = 1,
  sigma2_update_interval = 1,
  nu0_tol = 1e-4
)

data.frame(
  nu0 = fit_susie_relax_l2$nu0,
  tau2 = fit_susie_relax_l2$tau2,
  sigma2 = paste(sprintf("%.4g", fit_susie_relax_l2$sigma2), collapse = ", "),
  gamma_hat = paste(fit_susie_relax_l2$gamma_hat, collapse = ", "),
  top_alpha = paste(sprintf("%.3f", apply(fit_susie_relax_l2$alpha, 1, max)),
                    collapse = ", "),
  causal_total_pip = sum(fit_susie_relax_l2$pip[causal])
)
```

```{r plot-susie-relax-l2-pip}
plot_pips(
  list("SuSiE-relax\nL=2, estimated nu0" = fit_susie_relax_l2$pip),
  causal,
  beta_true
)
```

```{r plot-susie-relax-l2-convergence}
old_par <- par(no.readonly = TRUE)
on.exit(par(old_par), add = TRUE)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
elbo_iter <- which(!is.na(fit_susie_relax_l2$elbo))
plot(
  elbo_iter,
  fit_susie_relax_l2$elbo[elbo_iter],
  type = "b",
  pch = 16,
  lwd = 2,
  xlab = "Iteration",
  ylab = "ELBO",
  main = "SuSiE-relax L=2 ELBO"
)
grid()
plot(
  seq_along(fit_susie_relax_l2$nu0_trace),
  fit_susie_relax_l2$nu0_trace,
  type = "l",
  lwd = 2,
  xlab = "Iteration",
  ylab = expression(hat(nu)[0]),
  main = expression("SuSiE-relax L=2 " * hat(nu)[0])
)
grid()
par(old_par)
```
