---
title: "SER-relax PIP vs Model B PIP"
author: "Dat Do"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 4.8
)
```

# Goal

The goal is to compare the exact single-effect PIP weights from
SER-relax with the closed-form PIP weights from Model B in
`tex/susie-love-revisit-extension.tex`.

Model B has the closed-form weight
\[
w_j^B
\propto
N\left(x_j;0,\sigma_0^2+\frac1N\right)
\left(1+\frac{Nx_j^2}{\nu_0}\right)^{(\nu_0+3)/2}.
\]

For SER-relax, after integrating out \(z_j\), we get
\[
p(x\mid \gamma=j,\beta)
=
N\left(x_j;\beta,\frac1N\right)
N_{J-1}\left(
x_{-j};x_j\mu_j,
\left\{\frac1N+\frac{\beta^2}{\kappa_z}\right\}S_j
\right),
\]
Here we use the parameterization \(\kappa_z=\nu_0\), matching the
comparison calculation in `tex/susie-love-revisit-extension.tex`.

After integrating out \(\beta\),
\[
w_j^{\mathrm{relax}}
\propto
N\left(x_j;0,\sigma_0^2+\frac1N\right)
E_{\beta\mid x_j}
\left[
c(\beta)^{-(J-1)/2}
\exp\left\{-\frac{r_j^2}{2c(\beta)}\right\}
\right],
\]
where
\[
c(\beta)=\frac1N+\frac{\beta^2}{\kappa_z},
\qquad
r_j^2=x^\top\bar R^{-1}x-x_j^2,
\]
and
\[
\beta\mid x_j
\sim
N\left(
\frac{N}{N+\sigma_0^{-2}}x_j,
\frac{1}{N+\sigma_0^{-2}}
\right).
\]

The question is whether these normalized PIPs are actually close for
representative values of \(\nu_0\).

We also compare against the new shared-\(\omega\) model from
`tex/susie-love-revisit-extension.tex`. Its single-effect model is
\[
x\mid \beta,\gamma=j,z_j,\omega_j
\sim
N_J\left(\beta u_j(z_j),\frac{\bar R}{N\omega_j}\right),
\]
\[
z_j\mid\omega_j
\sim
N_{J-1}\left(\mu_j,\frac{S_j}{\nu_0\omega_j}\right),
\qquad
\omega_j\sim
\operatorname{Gamma}\left(\frac{\nu_0+3}{2},\frac{\nu_0}{2}\right).
\]
For this model we compute PIPs using a local mean-field VI update for
each candidate causal SNP.

# Helper Functions

```{r helpers}
ar_correlation <- function(J, rho) {
  idx <- seq_len(J)
  rho ^ abs(outer(idx, idx, "-"))
}

softmax <- function(logw) {
  z <- logw - max(logw)
  ez <- exp(z)
  ez / sum(ez)
}

logsumexp <- function(x) {
  m <- max(x)
  m + log(sum(exp(x - m)))
}

gauss_hermite <- function(n = 41) {
  i <- seq_len(n - 1)
  J <- matrix(0, n, n)
  off <- sqrt(i / 2)
  J[cbind(i, i + 1)] <- off
  J[cbind(i + 1, i)] <- off
  eg <- eigen(J, symmetric = TRUE)
  o <- order(eg$values)
  nodes <- eg$values[o]
  weights <- sqrt(pi) * eg$vectors[1, o]^2
  list(nodes = nodes, weights = weights)
}

model_b_pip <- function(x, N, sigma2, nu0) {
  logw <- dnorm(x, 0, sqrt(sigma2 + 1 / N), log = TRUE) +
    ((nu0 + 3) / 2) * log1p(N * x^2 / nu0)
  softmax(logw)
}

ser_relax_exact_pip <- function(x, Rbar, N, sigma2, nu0,
                                kappa_z = nu0,
                                gh = gauss_hermite(41)) {
  J <- length(x)
  p <- J - 1
  Omega_x <- solve(Rbar, x)
  q <- as.numeric(crossprod(x, Omega_x))
  r2 <- pmax(q - x^2, 0)

  v_beta <- 1 / (N + 1 / sigma2)
  m_beta <- (N / (N + 1 / sigma2)) * x
  sd_beta <- sqrt(v_beta)

  logw <- numeric(J)
  log_gh_w <- log(gh$weights / sqrt(pi))
  for (j in seq_len(J)) {
    beta_nodes <- m_beta[j] + sqrt(2) * sd_beta * gh$nodes
    c_nodes <- 1 / N + beta_nodes^2 / kappa_z
    log_kernel <- -(p / 2) * log(c_nodes) - r2[j] / (2 * c_nodes)
    log_expect <- logsumexp(log_gh_w + log_kernel)
    logw[j] <- dnorm(x[j], 0, sqrt(sigma2 + 1 / N), log = TRUE) +
      log_expect
  }
  softmax(logw)
}

shared_omega_vi_pip <- function(x, Rbar, N, sigma2, nu0,
                                max_iter = 500, tol = 1e-10) {
  J <- length(x)
  p <- J - 1
  Omega <- solve(Rbar)
  logdet_Rbar <- as.numeric(determinant(Rbar, logarithm = TRUE)$modulus)
  xOx <- as.numeric(crossprod(x, Omega %*% x))
  C <- pmax(xOx - x^2, 0)

  a0 <- (nu0 + 3) / 2
  b0 <- nu0 / 2
  a_omega <- a0 + (J + p) / 2

  compute_abq <- function(j, z_scale, z_denom) {
    Q <- p / z_denom + z_scale^2 * C[j]
    A <- 1 + Q
    B <- x[j] + z_scale * C[j]
    list(A = A, B = B, Q = Q)
  }

  local_elbo <- function(j, state) {
    eomega <- state$aomega / state$bomega
    elogomega <- digamma(state$aomega) - log(state$bomega)
    beta2 <- state$vbeta + state$mbeta^2
    abq <- compute_abq(j, state$z_scale, state$z_denom)
    Lquad <- xOx - 2 * state$mbeta * abq$B + beta2 * abq$A

    eloglik <- -J / 2 * log(2 * pi) -
      0.5 * (logdet_Rbar - J * log(N)) +
      J / 2 * elogomega -
      N / 2 * eomega * Lquad

    elog_pbeta <- -0.5 * log(2 * pi * sigma2) - beta2 / (2 * sigma2)

    elog_pz <- -p / 2 * log(2 * pi) -
      0.5 * logdet_Rbar +
      p / 2 * log(nu0) +
      p / 2 * elogomega -
      nu0 / 2 * eomega * abq$Q

    elog_pomega <- a0 * log(b0) - lgamma(a0) +
      (a0 - 1) * elogomega - b0 * eomega

    elog_qbeta <- -0.5 * log(2 * pi * state$vbeta) - 0.5
    logdet_vz <- logdet_Rbar - p * log(state$z_denom)
    elog_qz <- -p / 2 * log(2 * pi) - 0.5 * logdet_vz - p / 2
    elog_qomega <- state$aomega * log(state$bomega) -
      lgamma(state$aomega) +
      (state$aomega - 1) * elogomega -
      state$bomega * eomega

    -log(J) + eloglik + elog_pbeta + elog_pz + elog_pomega -
      elog_qbeta - elog_qz - elog_qomega
  }

  fit_one <- function(j) {
    eomega <- a0 / b0
    z_scale <- 0
    z_denom <- eomega * nu0
    state <- list(
      mbeta = 0,
      vbeta = sigma2,
      z_scale = z_scale,
      z_denom = z_denom,
      aomega = a_omega,
      bomega = b0 + 0.5 * N * xOx
    )

    old_elbo <- -Inf
    for (iter in seq_len(max_iter)) {
      eomega <- state$aomega / state$bomega

      abq <- compute_abq(j, state$z_scale, state$z_denom)
      vbeta <- 1 / (1 / sigma2 + N * eomega * abq$A)
      mbeta <- vbeta * N * eomega * abq$B
      beta2 <- vbeta + mbeta^2

      z_denom <- eomega * (N * beta2 + nu0)
      z_scale <- N * mbeta / (N * beta2 + nu0)

      abq <- compute_abq(j, z_scale, z_denom)
      Lquad <- xOx - 2 * mbeta * abq$B + beta2 * abq$A
      bomega <- b0 + N / 2 * Lquad + nu0 / 2 * abq$Q

      state <- list(
        mbeta = mbeta,
        vbeta = vbeta,
        z_scale = z_scale,
        z_denom = z_denom,
        aomega = a_omega,
        bomega = bomega
      )

      elbo <- local_elbo(j, state)
      if (is.finite(old_elbo) && abs(elbo - old_elbo) < tol * (1 + abs(old_elbo))) {
        break
      }
      old_elbo <- elbo
    }

    list(state = state, elbo = local_elbo(j, state))
  }

  fits <- lapply(seq_len(J), fit_one)
  local <- vapply(fits, `[[`, numeric(1), "elbo")
  list(
    pi = softmax(local),
    local_elbo = local,
    states = lapply(fits, `[[`, "state")
  )
}

pip_metrics <- function(pip, reference, label, nu0, causal) {
  data.frame(
    method = label,
    nu0 = nu0,
    top_snp = which.max(pip),
    top_pip = max(pip),
    causal_pip = sum(pip[causal]),
    max_abs_diff = max(abs(pip - reference)),
    total_variation = 0.5 * sum(abs(pip - reference)),
    corr_with_model_b = suppressWarnings(cor(pip, reference)),
    model_b_top_pip_at_method_top = reference[which.max(pip)]
  )
}

plot_compare <- function(pip_list, causal, main) {
  old_par <- par(no.readonly = TRUE)
  on.exit(par(old_par), add = TRUE)
  ymax <- max(unlist(pip_list))
  par(mfrow = c(length(pip_list), 1), mar = c(3, 4, 2, 1))
  for (nm in names(pip_list)) {
    pip <- pip_list[[nm]]
    plot(
      seq_along(pip), pip,
      type = "h",
      lwd = 1.2,
      ylim = c(0, ymax),
      xlab = "SNP index",
      ylab = "PIP",
      main = paste(main, nm)
    )
    points(causal, pip[causal], pch = 19, col = "red")
    rug(causal, col = "red", lwd = 2)
    grid()
  }
}
```

# Simulation Setup

This setup is deliberately simple and reproducible. We use an AR(1)
reference LD matrix so that the comparison only tests the PIP formulas,
not the coalescent simulator.

```{r simulate}
set.seed(1)

J <- 200L
N <- 20000L
rho <- 0.9
Rbar <- ar_correlation(J, rho)
causal <- 95L
beta_true <- 0.03
sigma2 <- beta_true^2

x <- as.numeric(beta_true * Rbar[, causal] +
  drop(t(chol(Rbar / N)) %*% rnorm(J)))

nu0_grid <- c(5, 10, 20, 50, 100, 500, 2000)

data.frame(
  J = J,
  N = N,
  rho = rho,
  causal = causal,
  beta_true = beta_true,
  sigma2 = sigma2,
  max_abs_x = max(abs(x)),
  top_x_snp = which.max(abs(x))
)
```

# PIP Comparison

```{r compare-pips}
gh <- gauss_hermite(51)

all_results <- list()
pip_examples <- list()

for (nu0 in nu0_grid) {
  pip_b <- model_b_pip(x, N, sigma2, nu0)

  pip_ser_extension <- ser_relax_exact_pip(
    x, Rbar, N, sigma2, nu0,
    kappa_z = nu0,
    gh = gh
  )

  fit_shared <- shared_omega_vi_pip(
    x, Rbar, N, sigma2, nu0,
    max_iter = 500,
    tol = 1e-10
  )
  pip_shared_vi <- fit_shared$pi

  all_results[[as.character(nu0)]] <- rbind(
    pip_metrics(pip_b, pip_b, "Model B closed form", nu0, causal),
    pip_metrics(pip_ser_extension, pip_b,
                "SER-relax exact, kappa_z=nu0", nu0, causal),
    pip_metrics(pip_shared_vi, pip_b,
                "shared-omega SER-relax VI", nu0, causal)
  )

  if (nu0 %in% c(10, 100, 2000)) {
    pip_examples[[paste0("nu0=", nu0, ", Model B")]] <- pip_b
    pip_examples[[paste0("nu0=", nu0, ", SER kappa=nu0")]] <-
      pip_ser_extension
    pip_examples[[paste0("nu0=", nu0, ", shared omega VI")]] <-
      pip_shared_vi
  }
}

comparison <- do.call(rbind, all_results)
row.names(comparison) <- NULL
comparison
```

```{r plot-selected}
plot_compare(
  pip_examples[c("nu0=10, Model B",
                 "nu0=10, SER kappa=nu0",
                 "nu0=10, shared omega VI")],
  causal,
  main = ""
)

plot_compare(
  pip_examples[c("nu0=100, Model B",
                 "nu0=100, SER kappa=nu0",
                 "nu0=100, shared omega VI")],
  causal,
  main = ""
)
```

# Difference as a Function of \(\nu_0\)

```{r plot-differences}
ser_rows <- comparison[comparison$method != "Model B closed form", ]

old_par <- par(no.readonly = TRUE)
on.exit(par(old_par), add = TRUE)
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

plot(
  NA,
  xlim = range(nu0_grid),
  ylim = range(ser_rows$total_variation),
  log = "x",
  xlab = expression(nu[0]),
  ylab = "Total variation distance",
  main = "PIP distance from Model B"
)
for (method in unique(ser_rows$method)) {
  rows <- ser_rows[ser_rows$method == method, ]
  lines(rows$nu0, rows$total_variation, type = "b", pch = 16)
}
legend("topright", legend = unique(ser_rows$method), lty = 1, pch = 16, bty = "n")
grid()

plot(
  NA,
  xlim = range(nu0_grid),
  ylim = range(ser_rows$max_abs_diff),
  log = "x",
  xlab = expression(nu[0]),
  ylab = "Maximum absolute PIP difference",
  main = "Worst SNP-wise difference"
)
for (method in unique(ser_rows$method)) {
  rows <- ser_rows[ser_rows$method == method, ]
  lines(rows$nu0, rows$max_abs_diff, type = "b", pch = 16)
}
legend("topright", legend = unique(ser_rows$method), lty = 1, pch = 16, bty = "n")
grid()
par(old_par)
```

# Takeaway

The comparison above answers the question empirically for this simple
single-effect simulation.

-   For large \(\nu_0\), all three PIPs become close because they reduce
    to the same leading SER-RSS-style weight.
-   For small or moderate \(\nu_0\), the original SER-relax exact PIP
    need not be close to Model B. The SER-relax exact weight still
    contains
    \[
    E_{\beta\mid x_j}
    \left[
    c(\beta)^{-(J-1)/2}
    \exp\left\{-\frac{r_j^2}{2c(\beta)}\right\}
    \right],
    \]
    so it can depend on the LD residual
    \(r_j^2=x^\top\bar R^{-1}x-x_j^2\). Model B's closed-form PIP
    cancels the corresponding \(x^\top\bar R^{-1}x\) term exactly and
    depends only on \(x_j^2\).
-   The shared-\(\omega\) model uses one scalar latent uncertainty in
    both the residual noise and the relaxed LD column. Its VI PIP is
    included to check whether this shared-scale construction moves the
    approximation closer to Model B in finite-\(\nu_0\) regimes. In this
    simulation it does: the shared-\(\omega\) total variation distance is
    much smaller than original SER-relax for every \(\nu_0\) in the grid.

So the answer is: original SER-relax is only asymptotically close to
Model B as \(\nu_0\) gets large, but the shared-\(\omega\) construction
is already a substantially better approximation at finite \(\nu_0\) in
this example.
