---
title: "SER-relax-prototype Demonstration"
author: "Dat Do"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6.5,
  fig.height = 4.5
)

source("code/SER-relax-prototype.R")
source("code/sim_2pop.R")
```

# Introduction

Last time we discuss how to learn $\nu_0$ from observing a single column
of the in-sample LD matrix $R$ from $IW(\nu_0 \bar R, \nu_0 + J + 1)$.
We already know that it works well for the noiseless case, and wonder
how well it works for the noisy case, where the observation is corrupted
by noise like in the susie model.

I attempt to write the noisy model as the following hierarchical model:

$$x | R \sim N_J(R_1, R / N), \quad R  \sim IW(\nu_0 \bar R, \nu_0 + J + 1), $$
where $R_1$ is the first column of $R$. So I was thinking how to make
inference from this model. Long story short, it all happened like this:

-   I quickly realized that making inference here is as hard as
    `susie-love`, due to the appearance of the latent $R$ in both the
    mean and the variance of the observation $x$.

-   I tried to make things easier, and notice that there is an
    approximation that works.

-   After a few more days of work, this finally leads to a nice model
    that I call `susie-relax` that approximates the original
    `susie-love` model.

The `susie-relax` model has the following properties:

(i) is more interpretable than `susie-love`;

(ii) has all updates existed in closed-form (including $\nu_0$);

(iii) runs faster and costs less storage;

(iv) stay conservative under high LD-mismatch and retain power under low
     LD-mismatch.

Now I proceed to introduce the approximation and the single-effect model
that naturally follows, named `susie-relax-prototype`. In the next
workflowr, I introduce a few more benign approximations that makes
this\~100 times faster while still being accurate. In the third
workflowr, I introduce the extension called `susie-relax` that is robust
under LD mismatch.

# The key approximation of the noise part

As mentioned, the previous model is hard to make inference because the
latent $R$ appears in both the mean and the noise part of $x$. Let's
make the following approximation:

$$x | R \sim N_J(R_1, R / N) \approx N_J(R_1, \bar R / N) $$ This
approximation is benign, since:

i.  Its error is of order $1 / (N * \nu_0)$

ii. If we only consider the noise part:
    $\epsilon | R \sim N_J(0, R / N), R\sim IW(\nu_0 \bar R, \nu_0 + J + 1)$,
    then marginally $\epsilon \sim t_{\nu_0 + 2}(0, \bar R / N)$, which
    approximate $N_J(0, \bar R / N)$ very well for $\nu_0$ as small as
    $30$.

Under this approximation, the model becomes:

$$x | R \sim N_J(R_1, \bar R / N),\quad R\sim IW(\nu_0 \bar R, \nu_0 + J + 1). $$
Making inference of $\nu_0$ under this model is not hard anymore,
because we understand the distribution of $R_1$ well and can utilize a
latent variable trick to make the model becomes normal-normal hierarchy.
Indeed, write $d_1 = R_{11}$ and
$z_1 = R_{-1, 1} / R_{11} = (R_{21}, \dots, R_{J1}) / R_{11} \in \mathbb{R}^{J-1}$,
then the IW distribution induces:

$$d_1 \sim \operatorname{IG}\left(\frac{\nu_0+2}{2}, \frac{\nu_0}{2}\right),
$$ and $$
z_1 \sim t_{\nu_0 + 3}\left(\mu_1, \dfrac{S_1}{\nu_0 + 3}\right),
$$

where

$$
\mu_1 = \bar R_{-1,1},
\qquad
S_1 = \bar R_{-1,-1} - \bar R_{-1, 1}\bar R_{1,-1}.
$$

This has a nice interpretation: $R_{1}$ is centered around $\bar R_1$,
and the covariance matrix is the Schur complement of $\bar R_{11}$
(which turns out to have a really nice computation advantage in
inference as we will see later).

Making inference with $z_1$ is easier by introducing latent variable
$\omega_1$:

$$z_1\mid \omega_1
\sim
N_{J-1}\left(\mu_1, \frac{S_1}{\omega_1}\right),
\qquad
\omega_1 \sim \operatorname{Gamma}\left(\frac{\nu_0+3}{2}, \frac12\right) = \chi_{\nu_0 + 3}^2.
$$

Hence, the distribution of $x | d_1, z_1$ is normal, and the
distribution of $z_1 | \omega_1$ is normal, which makes the inference
easy. A computation advantage that we mentioned is that the precision
matrix of the likelihood of $x | z_1$ is proportional to
$(\bar R^{-1})_{-1, -1} = S_1^{-1}$, which is also proportional to the
precision matrix of $z_1 | \omega_1$. Hence, in updating posterior
variance of $z_1$ only involves scalar computation, not involving any
matrix inversion!

We can make inferences of $\nu_0$ using VEB:

$$\max_{\nu_0, q} E_{q(z_1)q(d_1)q(\omega_1)} [\log p(x | d_1, z_1) + \log p(z_1 | \omega_1) - \log q(z_1)] - \text{KL}(q(d_1)\| p(d_1)) -  \text{KL}(q(\omega_1)\| p(\omega_1)).$$

Note that $\nu_0$ appears in $p(d_1)$ and $p(\omega_1)$. Updates of
$q(z_1)$ and $q(\omega_1)$ exist in closed-form, while $q(d_1)$ is not
but we can use quadrature approximation and it is fast since it is 1D.

# SER-relax-prototype Model

Knowing how to do inference with a single row case, we can actually
extend it to the single-effect regression model:

$$x | R, \beta, \gamma \sim N_J(\beta R_{\gamma}, \bar R / N)$$

where $R_{\gamma}$ is the $\gamma$-th column of $R$,

$$ R \sim IW(\nu_0 \bar R, \nu_0 + J + 1), $$ and

$$
\beta \sim N(0,\sigma_0^2),
\qquad
\gamma \sim \operatorname{Unif}\{1,\ldots,J\}.
$$

Similar to above, this model can also be inferred using VEB. I am
calling this `SER-relax-prototype` because if we change $R_{\gamma}$ to
$\bar R_{\gamma}$ in the first equation, we recover the susie-rss model.
The considered model allows each column $R_{j}$ to be a *relaxed*
version of $\bar R_j$, with the relaxed rule induced by IW distribution,
which turns out to have a nice computation advantage. The amount of
relax is controlled by $\nu_0$, as we saw in the distribution of $z_1$,
where $\nu_0 + 3$ plays the role of the precision scale of $z_1$.

# Inference Overview

**(might skip this in the first read and see the result of recovering**
$\nu_0$ **at the end)**

Let $x \in \mathbb R^J$ be the vector of marginal summary statistics and
let $\bar R$ be the reference LD correlation matrix. For a single active
effect, $$
x \mid \beta,\gamma=j,d_j,z_j
\sim
N_J\left(\beta d_j u_j(z_j), \frac{\bar R}{N}\right),
$$ where $u_j(z_j)$ is the vector with entry $j$ equal to $1$, and
entries $-j$ equal to $z_j$. Writing $$
d_j = R_{jj},
\qquad
z_j = R_{-j,j}/R_{jj},
$$ the inverse-Wishart prior implies $$
d_j \sim \operatorname{IG}\left(\frac{\nu_0+2}{2}, \frac{\nu_0}{2}\right),
$$ and $$
z_j \mid \omega_j
\sim
N_{J-1}\left(\mu_j, \frac{S_j}{\omega_j}\right),
\qquad
\omega_j \sim \operatorname{Gamma}\left(\frac{\nu_0+3}{2}, \frac12\right),
$$ where $$
\mu_j = \bar R_{-j,j},
\qquad
S_j = \bar R_{-j,-j} - \bar R_{-j,j}\bar R_{j,-j}.
$$

The variational approximation factorizes by candidate active coordinate:
$$
q(\gamma=j,\beta,d_j,z_j,\omega_j)
=
\pi_j q_j(\beta)q_j(d_j)q_j(z_j)q_j(\omega_j).
$$ We use $$
q_j(\beta) = N(m_{\beta j},v_{\beta j}),
\qquad
q_j(z_j) = N_{J-1}(m_{zj},V_{zj}),
$$ and $$
q_j(\omega_j) = \operatorname{Gamma}(\alpha_{\omega j}, \beta_{\omega j}).
$$ The factor $q_j(d_j)$ is non-standard and is optimized numerically.

Let $$
\bar \beta_j = m_{\beta j},
\qquad
\overline{\beta^2}_j = v_{\beta j} + m_{\beta j}^2,
\qquad
\bar d_j = E_j[d_j],
\qquad
\overline{d_j^2}=E_j[d_j^2],
\qquad
\bar\omega_j = E_j[\omega_j].
$$ Define $$
A_j =
1 + \operatorname{tr}(S_j^{-1}V_{zj})
+ (m_{zj}-\mu_j)'S_j^{-1}(m_{zj}-\mu_j),
$$ $$
B_j =
x_j + (m_{zj}-\mu_j)'S_j^{-1}(x_{-j}-\mu_jx_j),
$$ and $$
Q_j =
\operatorname{tr}(S_j^{-1}V_{zj})
+ (m_{zj}-\mu_j)'S_j^{-1}(m_{zj}-\mu_j).
$$

## Closed-form Updates and Interpretations

The update for the effect size is Gaussian: $$
v_{\beta j}
=
\left(\sigma_0^{-2}+N\overline{d_j^2}A_j\right)^{-1},
\qquad
m_{\beta j}
=
v_{\beta j}N\bar d_jB_j.
$$ Here $A_j$ is the relaxed-column second moment under the
$\bar R^{-1}$ geometry, and $B_j$ is the relaxed-column alignment with
the data.

Let $$
T_{1j} = \bar\beta_j\bar d_j,
\qquad
T_{2j} = \overline{\beta^2}_j\overline{d_j^2}.
$$ The update for $z_j$ is also Gaussian: $$
V_{zj}
=
\frac{S_j}{NT_{2j}+\bar\omega_j},
$$ $$
m_{zj}
=
\mu_j
+
\frac{NT_{1j}}{NT_{2j}+\bar\omega_j}
(x_{-j}-\mu_jx_j).
$$ Thus $z_j$ is shrunk toward the reference column mean $\mu_j$, and
the strength of shrinkage is controlled by $\bar\omega_j$, which in turn
depends on $\nu_0$.

The update for $\omega_j$ is Gamma: $$
\alpha_{\omega j} = \frac{\nu_0+J+2}{2},
\qquad
\beta_{\omega j} = \frac{1+Q_j}{2}.
$$ If the relaxed column deviates more from $\mu_j$, then $Q_j$ is
larger and the posterior mean of $\omega_j$ is smaller, allowing more
uncertainty in $z_j$.

The coordinate probabilities are updated by softmax: $$
\pi_j =
\frac{\exp(L_j)}
{\sum_{k=1}^J \exp(L_k)},
$$ where $L_j$ is the coordinate-specific ELBO.

## Learning $\nu_0$

Holding the variational distribution fixed, only the priors for $d_j$
and $\omega_j$ depend on $\nu_0$. Therefore we update $\nu_0$ by
maximizing $$
M(\nu_0)
=
\sum_{j=1}^J
\pi_j
\left\{
E_j\log p(d_j\mid \nu_0)
+E_j\log p(\omega_j\mid \nu_0)
\right\}.
$$ This objective is one-dimensional. The implementation optimizes over
$\log \nu_0$ using a bounded scalar optimizer. This is more stable than
unconstrained Newton updates because $\nu_0>0$ and the objective may be
flat for weak signals.

## Numerical Optimization of $q(d)$

The optimal factor $q_j(d_j)$ is not a named distribution: $$
q_j(d_j)
\propto
d_j^{-(\nu_0+2)/2-1}
\exp\left\{
-\frac{\nu_0}{2d_j}
-\frac{N}{2}\overline{\beta^2}_jA_jd_j^2
+N\bar\beta_jB_jd_j
\right\},
\qquad d_j>0.
$$ The implementation works on the log scale $r_j=\log d_j$. Including
the Jacobian, $$
\ell_j(r)
=
-\left(\frac{\nu_0+2}{2}+1\right)r
-\frac{\nu_0}{2e^r}
-\frac{N}{2}\overline{\beta^2}_jA_je^{2r}
+N\bar\beta_jB_je^r
+r.
$$ We find the mode of $\ell_j(r)$, place a grid around the mode, expand
the grid if the endpoints still have non-negligible mass, and compute
the needed moments by normalized quadrature: $$
E[d_j],\quad E[d_j^2],\quad E[d_j^{-1}],\quad E[\log d_j].
$$

# Grid Experiment: Recovering $\nu_0$

This experiment confirms that SER-relax-prototype can recover $\nu_0$,
which represents the amount of uncertainty of the columns of $R$ in
Single-effect regression.

The design is:

1.  Generate realistic in-sample and reference LD matrices $R$ and $R_0$
    using `sim_2pop()` with $N_{01}=N_0=2000$. The generated in-sample
    $R$ is not used below; this step is only used to get a realistic
    reference matrix $R_0$.
2.  Convert them to correlation matrices and set $$
    \bar R = 0.99R_0 + 0.01I.
    $$
3.  For each true $\nu_0$ in a grid, draw $$
    R^{\mathrm{true}} \sim IW_J(\nu_0\bar R,\nu_0+J+1),
    $$ generate a single-effect summary vector $x$, and estimate $\nu_0$
    using `fit_ser_relax_prototype()`.
4.  Plot $\hat\nu_0$ against the true $\nu_0$ on raw and log-log scales.

```{r experiment-helper}
sample_inverse_wishart <- function(df, scale) {
  W <- stats::rWishart(1, df = df, Sigma = solve(scale))[, , 1]
  solve(W)
}
```

```{r generate-rbar}
N = 20000L # sample size in the model
N01 <- 1L # for simulation of R only
N0 <- 2000L # for simulation of R only
J <- 500L

rbar_time <- system.time({
  coalescent <- sim_2pop(
    chrom = "chr22",
    start = 20e6,
    end = 21e6,
    T_split = 20,
    n_gwas = N01,
    n_ref = N0,
    J = J,
    h2 = 0.01,
    seed = 1
  )

  R0 <- cov2cor(coalescent$R0)
  Rbar <- 0.99 * R0 + 0.01 * diag(1, J)
  Rbar <- cov2cor(Rbar)
})

range(eigen(Rbar, symmetric = TRUE, only.values = TRUE)$values)
rbar_time
```

```{r run-grid}
nu0_grid <- c(25, 50, 100, 200, 400, 800, 1600)
n_rep <- 3
sigma0 <- 0.2
beta <- 0.2
gamma <- ceiling(J / 2)
nu0_init <- 500
nu0_bounds <- c(5, 5000)
max_iter <- 300
tol <- 1e-5
quad_grid_size <- 301
seed <- 100

out <- vector("list", length(nu0_grid) * n_rep)
i <- 0

for (nu0_true in nu0_grid) {
  for (rep_id in seq_len(n_rep)) {
    i <- i + 1
    sim_seed <- seed + i
    set.seed(sim_seed)

    iw_time <- system.time({
      R_true <- sample_inverse_wishart(
        df = nu0_true + J + 1,
        scale = nu0_true * Rbar
      )
    })

    mean_x <- beta * R_true[, gamma]

    x_time <- system.time({
      x <- as.numeric(MASS::mvrnorm(1, mu = mean_x, Sigma = Rbar / N))
    })

    fit_time <- system.time({
      fit <- fit_ser_relax_prototype(
        x = x,
        Rbar = Rbar,
        N = N,
        sigma0 = sigma0,
        nu0_init = nu0_init,
        estimate_nu0 = TRUE,
        max_iter = max_iter,
        tol = tol,
        nu0_bounds = nu0_bounds,
        quad_grid_size = quad_grid_size,
        quad_width = 10,
        verbose = FALSE
      )
    })

    out[[i]] <- data.frame(
      nu0_true = nu0_true,
      nu0_hat = fit$nu0,
      rep = rep_id,
      gamma_true = gamma,
      gamma_hat = fit$gamma_hat,
      top_pi = max(fit$pi),
      beta = beta,
      converged_iter = length(fit$elbo),
      iw_seconds = unname(iw_time["elapsed"]),
      x_seconds = unname(x_time["elapsed"]),
      fit_seconds = unname(fit_time["elapsed"])
    )
  }
}

nu0_results <- do.call(rbind, out)

nu0_results
```

```{r summarize-grid}
nu0_summary <- aggregate(
  cbind(nu0_hat, top_pi, converged_iter, iw_seconds, x_seconds, fit_seconds) ~ nu0_true,
  data = nu0_results,
  FUN = mean
)
nu0_summary

colSums(nu0_results[, c("iw_seconds", "x_seconds", "fit_seconds")])
```

```{r plot-raw}
plot(
  nu0_results$nu0_true,
  nu0_results$nu0_hat,
  pch = 19,
  col = "steelblue",
  xlab = expression("true " * nu[0]),
  ylab = expression("estimated " * nu[0]),
  main = expression("SER-relax-prototype " * nu[0] * " recovery")
)
abline(0, 1, lty = 2, col = "gray40")
lines(nu0_summary$nu0_true, nu0_summary$nu0_hat, type = "b", pch = 17, lwd = 2)
grid()
```

```{r plot-loglog}
plot(
  nu0_results$nu0_true,
  nu0_results$nu0_hat,
  log = "xy",
  pch = 19,
  col = "steelblue",
  xlab = expression("true " * nu[0] * " (log scale)"),
  ylab = expression("estimated " * nu[0] * " (log scale)"),
  main = expression("SER-relax-prototype " * nu[0] * " recovery, log-log")
)
abline(0, 1, lty = 2, col = "gray40")
lines(nu0_summary$nu0_true, nu0_summary$nu0_hat, type = "b", pch = 17, lwd = 2)
grid()
```
