---
title: "SER-relax Demonstration"
author: "Dat Do"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6.5,
  fig.height = 4.5
)

source("code/SER-relax.R")
source("code/sim_2pop.R")
```

# SER-relax Model

SER-relax is the simplified single-effect version of
`SER-relax-prototype`. It is derived from SER-relax-prototype by making
two additional approximations: $$
d_j = 1,
\qquad
\omega_j = \nu_0 + 3.
$$

These approximations are useful for two reasons.

-   First, $\bar R$ is a correlation matrix, so the diagonal of the
    target LD matrix is naturally close to one. The prototype variable
    $d_j$ mainly models diagonal scaling, but our scientific target is
    the uncertainty in the LD column shape, not the diagonal. Fixing
    $d_j=1$ removes the non-standard $q(d_j)$ quadrature step.
    Furthermore, we know that the in-sample $R$ is also a correlation
    matrix with diagonal 1. So this makes so much sense.

-   Second, the prototype writes the $t$-like prior on $z_j$ through a
    latent precision $\omega_j$. Replacing $\omega_j$ by its prior mean,
    $$
    E(\omega_j)=\nu_0+3,
    $$ keeps the main interpretation of $\nu_0$: larger $\nu_0$ shrinks
    the active LD column more strongly toward $\bar R_{\cdot j}$, while
    smaller $\nu_0$ allows more relaxation. On the other hand, we know
    that $\omega_j \sim \chi^2_{\nu_0 + 3}$ is well concentrated in
    $\nu_0 + 3 \pm \sqrt{\nu_0}$ anyway.

The SER-relax observation model is $$
x \mid \beta,\gamma=j,z_j
\sim
N_J\left(\beta u_j(z_j),\frac{\bar R}{N}\right),
$$ with $$
\beta \sim N(0,\sigma_0^2),
\qquad
\gamma \sim \operatorname{Unif}\{1,\ldots,J\}.
$$ $u_j(z_j)$ is the vector with entry $j$ equal to $1$, and entries
$-j$ equal to $z_j$, and
$$z_j \sim N_{J-1} \left(\mu_j, \dfrac{S_j}{\nu_0 + 3}\right) = N_{J-1}(\mu_j,\tau^2S_j). $$
where $$
\mu_j = \bar R_{-j,j},
\qquad
S_j = \bar R_{-j,-j} - \bar R_{-j,j}\bar R_{j,-j}.
$$

# Inference Overview

The variational approximation is $$
q(\gamma=j,\beta,z_j)
=
\pi_j q_j(\beta)q_j(z_j),
$$ where $$
q_j(\beta)=N(m_{\beta j},v_{\beta j}),
\qquad
q_j(z_j)=N_{J-1}(m_{zj},V_{zj}).
$$

Let $$
\bar\beta_j=m_{\beta j},
\qquad
\overline{\beta^2}_j=v_{\beta j}+m_{\beta j}^2.
$$ Define $$
A_j =
1 + \operatorname{tr}(S_j^{-1}V_{zj})
+(m_{zj}-\mu_j)'S_j^{-1}(m_{zj}-\mu_j),
$$ $$
B_j =
x_j+(m_{zj}-\mu_j)'S_j^{-1}(x_{-j}-\mu_jx_j),
$$ and $$
Q_j =
\operatorname{tr}(S_j^{-1}V_{zj})
+(m_{zj}-\mu_j)'S_j^{-1}(m_{zj}-\mu_j).
$$

## Closed-form Updates and Interpretations

The effect-size update is $$
v_{\beta j}
=
\left(\sigma_0^{-2}+NA_j\right)^{-1},
\qquad
m_{\beta j}
=
v_{\beta j}NB_j.
$$ The relaxed-column update is $$
V_{zj}
=
\frac{S_j}{N\overline{\beta^2}_j+\tau^{-2}},
$$ $$
m_{zj}
=
\mu_j+
\frac{N\bar\beta_j}
{N\overline{\beta^2}_j+\tau^{-2}}
(x_{-j}-\mu_jx_j).
$$ Thus the relaxed column mean is the reference column plus a residual
correction. The correction is larger when the effect is strong and the
data support moving away from the reference LD, and smaller when $\nu_0$
is large.

The coordinate probability is $$
\pi_j =
\frac{\exp(\mathcal L_j^{\mathrm{SER}})}
{\sum_{k=1}^J \exp(\mathcal L_k^{\mathrm{SER}})}.
$$

## Learning $\nu_0$

Holding $q$ fixed, the $\nu_0$-dependent part of the ELBO is $$
M(\nu_0)
=
\sum_{j=1}^J
\pi_j
\left\{
\frac{J-1}{2}\log(\nu_0+3)
-\frac{\nu_0+3}{2}Q_j
\right\}.
$$ Equivalently, in the $\tau^2$ parameterization, $$
\hat\tau^2
=
\frac{\sum_{j=1}^J\pi_jQ_j}{J-1},
\qquad
\hat\nu_0 = \frac{1}{\hat\tau^2}-3.
$$ The implementation applies lower and upper bounds to $\nu_0$. This
closed-form empirical-Bayes update is the main computational advantage
over the prototype.

## Implementation Algebra

Because $\bar R$ is a correlation matrix and $\Omega=\bar R^{-1}$, $$
(r_{-j}-\mu_jr_j)'S_j^{-1}(r_{-j}-\mu_jr_j)
=
r'\Omega r-r_j^2.
$$ Also, $$
|S_j|=|\bar R|.
$$ Thus the implementation avoids explicitly forming all $S_j^{-1}$
matrices. With $$
z_{\mathrm{denom},j}=N\overline{\beta^2}_j+\tau^{-2},
\qquad
z_{\mathrm{scale},j}=\frac{N\bar\beta_j}{z_{\mathrm{denom},j}},
$$ the required scalar quantities are $$
Q_j=\frac{J-1}{z_{\mathrm{denom},j}}+z_{\mathrm{scale},j}^2C_j,
\qquad
A_j=1+Q_j,
\qquad
B_j=x_j+z_{\mathrm{scale},j}C_j,
$$ where $C_j=x'\Omega x-x_j^2$.

# Grid Experiment: Recovering $\nu_0$

This experiment mirrors the SER-relax-prototype demonstration. The goal
is to show that SER-relax can still recover $\nu_0$, while running
faster because it avoids both $q(d_j)$ quadrature and the latent
$\omega_j$ update.

The design is:

1.  Generate a realistic reference LD matrix $R_0$ using `sim_2pop()`.
2.  Set $$
    \bar R = 0.99R_0 + 0.01I.
    $$
3.  For each true $\nu_0$, generate data exactly as in the prototype
    demonstration: $$
    R^{\mathrm{true}} \sim IW_J(\nu_0\bar R,\nu_0+J+1),
    \qquad
    x \sim N_J(\beta R^{\mathrm{true}}_{\cdot j}, \bar R/N).
    $$
4.  Estimate $\nu_0$ using `fit_ser_relax()`.
5.  Keep `fit_seconds` so the runtime can be compared directly with the
    prototype.

```{r experiment-helper}
sample_inverse_wishart <- function(df, scale) {
  W <- stats::rWishart(1, df = df, Sigma = solve(scale))[, , 1]
  solve(W)
}
```

```{r generate-rbar}
N = 20000L # sample size in the model
N01 <- 1L # for simulation of R only
N0 <- 2000L # for simulation of R only
J <- 500L

rbar_time <- system.time({
  coalescent <- sim_2pop(
    chrom = "chr22",
    start = 20e6,
    end = 21e6,
    T_split = 20,
    n_gwas = N01,
    n_ref = N0,
    J = J,
    h2 = 0.01,
    seed = 1
  )

  R0 <- cov2cor(coalescent$R0)
  Rbar <- 0.99 * R0 + 0.01 * diag(1, J)
  Rbar <- cov2cor(Rbar)
})

range(eigen(Rbar, symmetric = TRUE, only.values = TRUE)$values)
rbar_time
```

Here we generate $x$ exactly as in the prototype file to show that the
approximation inference is very benign, while being much faster. We
should look at the `fit_second` and compare it with the
`SER-relax-prototype` file.

```{r run-grid}
nu0_grid <- c(25, 50, 100, 200, 400, 800, 1600)
n_rep <- 3
sigma0 <- 0.2
beta <- 0.2
gamma <- ceiling(J / 2)
nu0_init <- 500
nu0_bounds <- c(5, 5000)
max_iter <- 300
tol <- 1e-5
seed <- 100

out <- vector("list", length(nu0_grid) * n_rep)
i <- 0

for (nu0_true in nu0_grid) {
  for (rep_id in seq_len(n_rep)) {
    i <- i + 1
    sim_seed <- seed + i
    set.seed(sim_seed)

    iw_time <- system.time({
      R_true <- sample_inverse_wishart(
        df = nu0_true + J + 1,
        scale = nu0_true * Rbar
      )
    })

    mean_x <- beta * R_true[, gamma]

    x_time <- system.time({
      x <- as.numeric(MASS::mvrnorm(1, mu = mean_x, Sigma = Rbar / N))
    })

    fit_time <- system.time({
      fit <- fit_ser_relax(
        x = x,
        Rbar = Rbar,
        N = N,
        sigma0 = sigma0,
        nu0_init = nu0_init,
        estimate_nu0 = TRUE,
        max_iter = max_iter,
        tol = tol,
        nu0_bounds = nu0_bounds,
        verbose = FALSE
      )
    })

    out[[i]] <- data.frame(
      nu0_true = nu0_true,
      nu0_hat = fit$nu0,
      rep = rep_id,
      gamma_true = gamma,
      gamma_hat = fit$gamma_hat,
      top_pi = max(fit$pi),
      beta = beta,
      converged_iter = length(fit$elbo),
      iw_seconds = unname(iw_time["elapsed"]),
      x_seconds = unname(x_time["elapsed"]),
      fit_seconds = unname(fit_time["elapsed"])
    )
  }
}

nu0_results <- do.call(rbind, out)

nu0_results
```

```{r summarize-grid}
nu0_summary <- aggregate(
  cbind(nu0_hat, top_pi, converged_iter, iw_seconds, x_seconds, fit_seconds) ~ nu0_true,
  data = nu0_results,
  FUN = mean
)
nu0_summary

colSums(nu0_results[, c("iw_seconds", "x_seconds", "fit_seconds")])
```

```{r plot-raw}
plot(
  nu0_results$nu0_true,
  nu0_results$nu0_hat,
  pch = 19,
  col = "steelblue",
  xlab = expression("true " * nu[0]),
  ylab = expression("estimated " * nu[0]),
  main = expression("SER-relax " * nu[0] * " recovery")
)
abline(0, 1, lty = 2, col = "gray40")
lines(nu0_summary$nu0_true, nu0_summary$nu0_hat, type = "b", pch = 17, lwd = 2)
grid()
```

```{r plot-loglog}
plot(
  nu0_results$nu0_true,
  nu0_results$nu0_hat,
  log = "xy",
  pch = 19,
  col = "steelblue",
  xlab = expression("true " * nu[0] * " (log scale)"),
  ylab = expression("estimated " * nu[0] * " (log scale)"),
  main = expression("SER-relax " * nu[0] * " recovery, log-log")
)
abline(0, 1, lty = 2, col = "gray40")
lines(nu0_summary$nu0_true, nu0_summary$nu0_hat, type = "b", pch = 17, lwd = 2)
grid()
```
