---
title: "SER-collapse-R Demonstration"
author: "Dat Do"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6.5,
  fig.height = 4.5
)

source("code/SER-collapse-R.R")
source("code/sim_2pop.R")
source("code/plot.R")

plot_pip <- function(fit, R_for_cs, beta_true, main) {
  susie_plot_alpha(
    fit,
    R = R_for_cs,
    b = beta_true,
    main = main,
    coverage = 0.95,
    min_abs_corr = 0.5
  )
  grid()
}

as_susie_single_effect <- function(fit) {
  list(
    alpha = matrix(fit$pi, nrow = 1),
    pip = as.numeric(fit$pi)
  )
}
```

# Overview

This page tests the collapsed-\(R\) single-effect approach derived in
`tex/collapse-R-approach.tex`. The model integrates out the latent LD
matrix \(R\) under
$$
R \sim IW_J(\nu_0\bar R,\nu_0+J+1)
$$
and then fits a single-effect model
$$
b=\beta e_\gamma,
\qquad
\beta\sim N(0,\sigma_0^2).
$$

For a candidate causal SNP \(j\), define
$$
a_j(\nu_0)=\nu_0\bar R_{jj}+Nx_j^2.
$$
The collapsed likelihood contains the one-dimensional expectation
$$
E_{\omega\sim\chi^2_{\nu_0+3}}
\left[
\exp\left\{
-\frac{N a_j(\nu_0)\beta^2}{2\omega}
\right\}
\right].
$$

The code implements three approximations to this term:

-   `local`: use one local variational distribution \(q_j(\omega)\) for
    each candidate SNP.
-   `shared`: use one global variational distribution \(q(\omega)\)
    shared across all candidate SNPs.
-   `plugin`: replace \(\omega\) by its prior mean \(\nu_0+3\).

For each candidate value of \(\nu_0\), the algorithm optimizes the
corresponding coordinate updates and then chooses the \(\nu_0\) value
with the largest optimized lower bound or plug-in objective.

# Simulation Design

This mirrors `analysis/SER-relax-demonstrate.Rmd`, except that the
fitting function is `fit_ser_collapse_r()`.

1.  Generate a realistic reference LD matrix \(R_0\) using `sim_2pop()`.
2.  Set
    $$
    \bar R = 0.99R_0+0.01I.
    $$
3.  For each true \(\nu_0\), generate
    $$
    R^{\mathrm{true}}\sim IW_J(\nu_0\bar R,\nu_0+J+1),
    \qquad
    x\sim N_J(\beta R^{\mathrm{true}}_{\cdot j},\bar R/N).
    $$
4.  Fit the collapsed-\(R\) single-effect model under the three
    variants.

The summary statistic is simulated with one true causal SNP, located at
the middle SNP in the region.

```{r experiment-settings}
nu0_grid <- c(50, 200, 1000)
variants <- c("local", "shared", "plugin")

N <- 20000L
N01 <- 1L
N0 <- 2000L
J <- 500L

n_rep <- 3
sigma0 <- 0.2
beta <- 0.2
gamma <- ceiling(J / 2)
max_iter <- 300
tol <- 1e-5
seed <- 100

fit_nu0_grid <- sort(unique(c(
  exp(seq(log(5), log(5000), length.out = 41)),
  nu0_grid,
  10
)))
```

```{r experiment-helper}
run_ser_collapse_r_recovery_demo <- function(
    nu0_grid,
    variant,
    n_rep,
    N,
    J,
    sigma0,
    beta,
    gamma,
    fit_nu0_grid,
    max_iter,
    tol,
    seed,
    Rbar,
    verbose = FALSE) {
  out <- vector("list", length(nu0_grid) * n_rep)
  i <- 0

  for (nu0_true in nu0_grid) {
    for (rep_id in seq_len(n_rep)) {
      i <- i + 1
      sim_seed <- seed + i
      set.seed(sim_seed)

      iw_time <- system.time({
        R_true <- sample_inverse_wishart(
          df = nu0_true + J + 1,
          scale = nu0_true * Rbar
        )
      })
      ## standardize this: this does not match the model but it is more realistic for real application
      # R_true = cov2cor(R_true)
      mean_x <- beta * R_true[, gamma]

      x_time <- system.time({
        x <- as.numeric(MASS::mvrnorm(1, mu = mean_x, Sigma = Rbar / N))
      })

      fit_time <- system.time({
        fit <- fit_ser_collapse_r(
          x = x,
          Rbar = Rbar,
          N = N,
          sigma0 = sigma0,
          nu0_grid = fit_nu0_grid,
          variant = variant,
          max_iter = max_iter,
          tol = tol,
          verbose = verbose
        )
      })

      out[[i]] <- data.frame(
        nu0_true = nu0_true,
        nu0_hat = fit$nu0,
        rep = rep_id,
        gamma_true = gamma,
        gamma_hat = unname(fit$gamma_hat),
        top_pi = unname(max(fit$pi)),
        beta = beta,
        variant = variant,
        converged_iter = length(fit$elbo),
        iw_seconds = unname(iw_time["elapsed"]),
        x_seconds = unname(x_time["elapsed"]),
        fit_seconds = unname(fit_time["elapsed"])
      )
    }
  }

  results <- do.call(rbind, out)
  list(
    results = results,
    fit_nu0_grid = fit_nu0_grid
  )
}
```

# Generate Reference LD

```{r generate-rbar}
rbar_time <- system.time({
  coalescent <- sim_2pop(
    chrom = "chr22",
    start = 20e6,
    end = 21e6,
    T_split = 20,
    n_gwas = N01,
    n_ref = N0,
    J = J,
    h2 = 0.01,
    seed = 1
  )

  R0 <- cov2cor(coalescent$R0)
  Rbar <- 0.99 * R0 + 0.01 * diag(1, J)
  Rbar <- cov2cor(Rbar)
})

range(eigen(Rbar, symmetric = TRUE, only.values = TRUE)$values)
rbar_time
```

# Run Recovery Grid

The same \(R^{\mathrm{true}}\) and \(x\) draws are used for all three
variants because the same `Rbar` and simulation seeds are reused.

```{r run-grid}
variant_runs <- vector("list", length(variants))
names(variant_runs) <- variants

for (variant in variants) {
  variant_runs[[variant]] <- run_ser_collapse_r_recovery_demo(
    nu0_grid = nu0_grid,
    variant = variant,
    n_rep = n_rep,
    N = N,
    J = J,
    sigma0 = sigma0,
    beta = beta,
    gamma = gamma,
    fit_nu0_grid = fit_nu0_grid,
    max_iter = max_iter,
    tol = tol,
    seed = seed,
    Rbar = Rbar,
    verbose = FALSE
  )
}

nu0_results <- do.call(rbind, lapply(variant_runs, `[[`, "results"))
row.names(nu0_results) <- NULL

nu0_results
```

# Summaries

```{r summarize-grid}
nu0_summary <- aggregate(
  cbind(nu0_hat, top_pi, converged_iter, iw_seconds, x_seconds, fit_seconds) ~
    variant + nu0_true,
  data = nu0_results,
  FUN = mean
)

nu0_summary
```

```{r timing}
aggregate(
  cbind(iw_seconds, x_seconds, fit_seconds) ~ variant,
  data = nu0_results,
  FUN = sum
)
```

# Plots

```{r plot-raw}
variant_cols <- c(local = "steelblue", shared = "darkorange", plugin = "firebrick")
variant_pch <- c(local = 19, shared = 17, plugin = 15)

plot(
  nu0_results$nu0_true,
  nu0_results$nu0_hat,
  type = "n",
  xlab = expression("true " * nu[0]),
  ylab = expression("estimated " * nu[0]),
  main = expression("SER-collapse-R " * nu[0] * " recovery")
)
abline(0, 1, lty = 2, col = "gray40")

for (variant in variants) {
  idx <- nu0_results$variant == variant
  points(
    nu0_results$nu0_true[idx],
    nu0_results$nu0_hat[idx],
    pch = variant_pch[variant],
    col = variant_cols[variant]
  )

  sid <- nu0_summary$variant == variant
  lines(
    nu0_summary$nu0_true[sid],
    nu0_summary$nu0_hat[sid],
    type = "b",
    pch = variant_pch[variant],
    col = variant_cols[variant],
    lwd = 2
  )
}

legend(
  "topleft",
  legend = variants,
  col = variant_cols[variants],
  pch = variant_pch[variants],
  lwd = 2,
  bty = "n"
)
grid()
```

```{r plot-loglog}
plot(
  nu0_results$nu0_true,
  nu0_results$nu0_hat,
  log = "xy",
  type = "n",
  xlab = expression("true " * nu[0] * " (log scale)"),
  ylab = expression("estimated " * nu[0] * " (log scale)"),
  main = expression("SER-collapse-R " * nu[0] * " recovery, log-log")
)
abline(0, 1, lty = 2, col = "gray40")

for (variant in variants) {
  idx <- nu0_results$variant == variant
  points(
    nu0_results$nu0_true[idx],
    nu0_results$nu0_hat[idx],
    pch = variant_pch[variant],
    col = variant_cols[variant]
  )

  sid <- nu0_summary$variant == variant
  lines(
    nu0_summary$nu0_true[sid],
    nu0_summary$nu0_hat[sid],
    type = "b",
    pch = variant_pch[variant],
    col = variant_cols[variant],
    lwd = 2
  )
}

legend(
  "topleft",
  legend = variants,
  col = variant_cols[variants],
  pch = variant_pch[variants],
  lwd = 2,
  bty = "n"
)
grid()
```

# PIP Plots

Here are representative replicates for \(\nu_0=200\) and
\(\nu_0=10\). For each case, the two plots show the PIPs from the
local-\(q_j(\omega)\) and shared-\(q(\omega)\) collapsed-\(R\) fits on
the same simulated data set.

```{r pip-example-fit}
fit_pip_example <- function(nu0_true, sim_seed) {
  set.seed(sim_seed)
  R_true <- sample_inverse_wishart(
    df = nu0_true + J + 1,
    scale = nu0_true * Rbar
  )
  x <- as.numeric(MASS::mvrnorm(
    1,
    mu = beta * R_true[, gamma],
    Sigma = Rbar / N
  ))

  beta_true <- rep(0, J)
  beta_true[gamma] <- beta

  fit_local <- fit_ser_collapse_r(
    x = x,
    Rbar = Rbar,
    N = N,
    sigma0 = sigma0,
    nu0_grid = fit_nu0_grid,
    variant = "local",
    max_iter = max_iter,
    tol = tol,
    verbose = FALSE
  )

  fit_shared <- fit_ser_collapse_r(
    x = x,
    Rbar = Rbar,
    N = N,
    sigma0 = sigma0,
    nu0_grid = fit_nu0_grid,
    variant = "shared",
    max_iter = max_iter,
    tol = tol,
    verbose = FALSE
  )

  list(
    beta_true = beta_true,
    local = fit_local,
    shared = fit_shared
  )
}

pip_nu0_200 <- fit_pip_example(nu0_true = 200, sim_seed = seed + 4)
pip_nu0_10 <- fit_pip_example(nu0_true = 10, sim_seed = seed + 40)

rbind(
  data.frame(
    true_nu0 = 200,
    variant = c("local", "shared"),
    nu0_hat = c(pip_nu0_200$local$nu0, pip_nu0_200$shared$nu0),
    gamma_hat = c(pip_nu0_200$local$gamma_hat, pip_nu0_200$shared$gamma_hat),
    top_pi = c(max(pip_nu0_200$local$pi), max(pip_nu0_200$shared$pi))
  ),
  data.frame(
    true_nu0 = 10,
    variant = c("local", "shared"),
    nu0_hat = c(pip_nu0_10$local$nu0, pip_nu0_10$shared$nu0),
    gamma_hat = c(pip_nu0_10$local$gamma_hat, pip_nu0_10$shared$gamma_hat),
    top_pi = c(max(pip_nu0_10$local$pi), max(pip_nu0_10$shared$pi))
  )
)
```

```{r plot-pip-local-nu0-200}
plot_pip(
  as_susie_single_effect(pip_nu0_200$local),
  R_for_cs = Rbar,
  beta_true = pip_nu0_200$beta_true,
  main = "Collapsed-R local q_j(omega) PIP, true nu0 = 200"
)
```

```{r plot-pip-shared-nu0-200}
plot_pip(
  as_susie_single_effect(pip_nu0_200$shared),
  R_for_cs = Rbar,
  beta_true = pip_nu0_200$beta_true,
  main = "Collapsed-R shared q(omega) PIP, true nu0 = 200"
)
```

```{r plot-pip-local-nu0-10}
plot_pip(
  as_susie_single_effect(pip_nu0_10$local),
  R_for_cs = Rbar,
  beta_true = pip_nu0_10$beta_true,
  main = "Collapsed-R local q_j(omega) PIP, true nu0 = 10"
)
```

```{r plot-pip-shared-nu0-10}
plot_pip(
  as_susie_single_effect(pip_nu0_10$shared),
  R_for_cs = Rbar,
  beta_true = pip_nu0_10$beta_true,
  main = "Collapsed-R shared q(omega) PIP, true nu0 = 10"
)
```
