---
title: "SuSiE-relax Demonstration"
author: "Dat Do"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)

library(susieR)

source("code/SuSiE-relax.R")
source("code/sim_2pop.R")
source("code/plot.R")
```

# From SER-relax to SuSiE-relax

SER-relax models one active effect while allowing the active LD column to relax around the reference LD matrix. SuSiE-relax extends this idea to \(L\) possible single effects:
\[
x \mid \{\beta_\ell,\gamma_\ell,z_{\ell,\gamma_\ell}\}_{\ell=1}^L
\sim
N_J\left(
\sum_{\ell=1}^L
\beta_\ell u_{\gamma_\ell}(z_{\ell,\gamma_\ell}),
\frac{\bar R}{N}
\right).
\]
For each effect,
\[
\gamma_\ell \sim \operatorname{Unif}\{1,\ldots,J\},
\qquad
\beta_\ell \sim N(0,\sigma_\ell^2),
\]
and, conditional on \(\gamma_\ell=j\),
\[
z_{\ell j}\mid \nu_0
\sim
N_{J-1}\left(\mu_j,\frac{S_j}{\nu_0+3}\right).
\]
Here
\[
\mu_j=\bar R_{-j,j},
\qquad
S_j=\bar R_{-j,-j}-\bar R_{-j,j}\bar R_{j,-j}.
\]

The variational approximation factorizes over effects:
\[
q =
\prod_{\ell=1}^L
q_\ell(\gamma_\ell)
\prod_{j=1}^J
q_{\ell j}(\beta_\ell)q_{\ell j}(z_{\ell j}).
\]
The inference algorithm is IBSS-like. When updating effect \(\ell\), we form the residual
\[
r_\ell = x-\sum_{k\ne\ell}\bar\theta_k,
\]
where
\[
\bar\theta_\ell =
\sum_{j=1}^J
\pi_{\ell j}E\{\beta_\ell u_j(z_{\ell j})\}.
\]
Then we apply the SER-relax update to this residual. After each full sweep over effects, the shared \(\nu_0\) is updated by the closed-form empirical-Bayes update
\[
\hat\tau^2 =
\frac{
\sum_{\ell=1}^L\sum_{j=1}^J\pi_{\ell j}Q_{\ell j}
}{L(J-1)},
\qquad
\hat\nu_0 = \frac{1}{\hat\tau^2}-3,
\]
with lower and upper bounds in implementation.

The main empirical behavior we want to demonstrate is:

- When the reference LD mismatch is low, ordinary SuSiE and SuSiE-relax both recover the two true signals.

- When the reference LD mismatch is high, ordinary SuSiE with out-of-sample LD can overfit and select all \(L=5\) effects, while SuSiE-relax responds by learning a small \(\nu_0\), using stronger LD relaxation, and selecting only one effect.

```{r helper-functions}
standardize_sim <- function(res) {
  d <- 1 / sqrt(diag(res$R))
  res$v <- d * res$v
  res$R <- cov2cor(res$R)
  res$R0 <- cov2cor(res$R0)
  res
}

make_rbar <- function(R0, lambda = 0.01) {
  Rbar <- (1 - lambda) * R0 + lambda * diag(1, ncol(R0))
  cov2cor(Rbar)
}

make_cs_from_alpha <- function(alpha, R, coverage = 0.95,
                               min_abs_corr = 0.5,
                               min_top_alpha = 1e-4) {
  alpha <- as.matrix(alpha)
  cs <- list()
  purity <- numeric()

  for (ell in seq_len(nrow(alpha))) {
    a <- as.numeric(alpha[ell, ])
    if (max(a) < min_top_alpha) {
      next
    }

    ord <- order(a, decreasing = TRUE)
    k <- which(cumsum(a[ord]) >= coverage)[1]
    vars <- ord[seq_len(k)]
    p <- if (length(vars) == 1) {
      1
    } else {
      R_sub <- abs(R[vars, vars, drop = FALSE])
      min(R_sub[upper.tri(R_sub)])
    }

    if (is.finite(p) && p >= min_abs_corr) {
      nm <- paste0("L", ell)
      cs[[nm]] <- vars
      purity[nm] <- p
    }
  }

  list(cs = cs, purity = purity)
}

count_selected_cs <- function(alpha, R, coverage = 0.95,
                              min_abs_corr = 0.5,
                              min_top_alpha = 1e-4) {
  length(make_cs_from_alpha(
    alpha = alpha,
    R = R,
    coverage = coverage,
    min_abs_corr = min_abs_corr,
    min_top_alpha = min_top_alpha
  )$cs)
}

summarize_fit <- function(label, alpha, pip, R_for_cs, beta_true,
                          nu0 = NA_real_) {
  alpha <- as.matrix(alpha)
  top <- apply(alpha, 1, which.max)
  top_alpha <- apply(alpha, 1, max)
  cs_info <- make_cs_from_alpha(alpha, R = R_for_cs)

  data.frame(
    method = label,
    selected_L = length(cs_info$cs),
    nu0 = nu0,
    gamma_hat = paste(top, collapse = ", "),
    top_alpha = paste(sprintf("%.3f", top_alpha), collapse = ", "),
    causal_pip = paste(sprintf("%.3f", pip[beta_true != 0]), collapse = ", ")
  )
}

plot_pip <- function(fit, R_for_cs, beta_true, main) {
  susie_plot_alpha(
    fit,
    R = R_for_cs,
    b = beta_true,
    main = main,
    coverage = 0.95,
    min_abs_corr = 0.5
  )
  grid()
}

plot_convergence <- function(fit, main_prefix) {
  old_par <- par(no.readonly = TRUE)
  on.exit(par(old_par), add = TRUE)

  par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
  plot(seq_along(fit$elbo), fit$elbo, type = "l", lwd = 2,
       xlab = "Iteration", ylab = "ELBO",
       main = paste(main_prefix, "ELBO"))
  grid()
  plot(seq_along(fit$nu0_trace), fit$nu0_trace, type = "l", lwd = 2,
       xlab = "Iteration", ylab = expression(hat(nu)[0]),
       main = paste(main_prefix, expression(nu[0])))
  grid()
}

plot_tau2_convergence <- function(fit, main_prefix) {
  tau2_trace <- 1 / (fit$nu0_trace + 3)
  plot(
    seq_along(tau2_trace), tau2_trace,
    type = "l",
    lwd = 2,
    xlab = "Iteration",
    ylab = expression(1 / (hat(nu)[0] + 3)),
    main = paste(main_prefix, expression(1 / (nu[0] + 3)))
  )
  grid()
}

run_case <- function(case_label, N0, seed, N = 20000L, J = 1000L,
                     lambda = 0.01, L = 5,
                     relax_max_iter = 3000,
                     relax_nu0_init = 2000,
                     relax_nu0_bounds = c(10, 10000)) {
  res <- sim_2pop(
    chrom = "chr22",
    start = 20e6,
    end = 21e6,
    T_split = 20,
    n_gwas = N,
    n_ref = N0,
    J = J,
    h2 = 0.01,
    seed = seed
  )
  res <- standardize_sim(res)

  v <- res$v
  R <- res$R
  R0 <- res$R0
  Rbar <- make_rbar(R0, lambda = lambda)
  beta_true <- res$beta_true

  fit_susie_out <- susie_suff_stat(
    XtX = N * R0,
    Xty = N * v,
    n = N,
    yty = N,
    L = L
  )

  relax_fit_time <- system.time({
    fit_relax <- fit_susie_relax(
      x = v,
      Rbar = Rbar,
      N = N,
      L = L,
      sigma2 = rep(0.3^2, L),
      nu0_init = relax_nu0_init,
      estimate_nu0 = TRUE,
      warmup_iter = 5,
      max_iter = relax_max_iter,
      tol = 1e-7,
      verbose = FALSE,
      nu0_bounds = relax_nu0_bounds,
      nu0_update_interval = 1,
      nu0_tol = 1e-4
    )
  })

  summary <- rbind(
    summarize_fit(
      label = "susieR with R0",
      alpha = fit_susie_out$alpha,
      pip = fit_susie_out$pip,
      R_for_cs = R0,
      beta_true = beta_true
    ),
    summarize_fit(
      label = "SuSiE-relax",
      alpha = fit_relax$alpha,
      pip = fit_relax$pip,
      R_for_cs = Rbar,
      beta_true = beta_true,
      nu0 = fit_relax$nu0
    )
  )

  list(
    case_label = case_label,
    N = N,
    N0 = N0,
    seed = seed,
    lambda = lambda,
    L = L,
    v = v,
    R = R,
    R0 = R0,
    Rbar = Rbar,
    beta_true = beta_true,
    causal = which(beta_true != 0),
    fit_susie_out = fit_susie_out,
    fit_relax = fit_relax,
    relax_fit_time = relax_fit_time,
    summary = summary
  )
}
```

# Low-mismatch Case

Here the reference panel size is \(N_0=1000\), so the out-of-sample LD is close enough to the GWAS LD. We expect both ordinary SuSiE and SuSiE-relax to select two effects.

```{r low-mismatch-fit}
low_case <- run_case(
  case_label = "low mismatch",
  N0 = 2000L,
  seed = 6,
  N = 20000L,
  J = 1000L,
  lambda = 0.01,
  L = 5,
  relax_max_iter = 1000,
  relax_nu0_init = 2000,
  relax_nu0_bounds = c(10, 10000)
)

low_case$relax_fit_time
low_case$causal
low_case$summary
```

```{r low-mismatch-plots}
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
plot_pip(
  low_case$fit_susie_out,
  low_case$R0,
  low_case$beta_true,
  main = "Low mismatch: susieR with R0"
)
plot_pip(
  low_case$fit_relax,
  low_case$Rbar,
  low_case$beta_true,
  main = "Low mismatch: SuSiE-relax"
)
par(old_par)

plot_convergence(low_case$fit_relax, "Low mismatch SuSiE-relax")
```

It might make more sense to look at \(1/(\nu_0+3)\) instead of \(\nu_0\), because this is the relaxed-column variance scale used by the model. Large changes in \(\nu_0\) at high values may correspond to very small changes in the actual relaxation variance.

```{r low-mismatch-tau2-convergence}
plot_tau2_convergence(low_case$fit_relax, "Low mismatch SuSiE-relax")
```

# High-mismatch Case

Here the reference panel size is \(N_0=100\). The out-of-sample LD is much noisier. We fit ordinary SuSiE with upper bound \(L=5\) using the out-of-sample LD \(R_0\), and fit SuSiE-relax with the stabilized matrix
\[
\bar R = 0.99R_0 + 0.01I.
\]
The expected behavior is that ordinary SuSiE overfits and selects all five effects, while SuSiE-relax learns a small \(\nu_0\), relaxes the LD column, and selects only one effect.

```{r high-mismatch-fit}
high_case <- run_case(
  case_label = "high mismatch",
  N0 = 100L,
  seed = 9,
  N = 20000L,
  J = 1000L,
  lambda = 0.01,
  L = 5,
  relax_max_iter = 3000,
  relax_nu0_init = 2000,
  relax_nu0_bounds = c(10, 10000)
)

high_case$relax_fit_time
high_case$causal
high_case$summary
```

```{r high-mismatch-plots}
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
plot_pip(
  high_case$fit_susie_out,
  high_case$R0,
  high_case$beta_true,
  main = "High mismatch: susieR with R0"
)
plot_pip(
  high_case$fit_relax,
  high_case$Rbar,
  high_case$beta_true,
  main = "High mismatch: SuSiE-relax"
)
par(old_par)

plot_convergence(high_case$fit_relax, "High mismatch SuSiE-relax")
```

```{r high-mismatch-effect-details}
data.frame(
  effect = seq_len(nrow(high_case$fit_relax$alpha)),
  top = apply(high_case$fit_relax$alpha, 1, which.max),
  top_alpha = apply(high_case$fit_relax$alpha, 1, max)
)
```
