---
title: "SuSiE-relax-prime Demonstration"
author: "Dat Do"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)

library(susieR)

source("code/susie-relax-prime.R")
source("code/sim_2pop.R")
source("code/plot.R")
```

# From SER-relax-prime to SuSiE-relax-prime

This page mirrors `susie-relax-demonstrate.Rmd`, but replaces the original
SuSiE-relax fit with the prime implementation in `code/susie-relax-prime.R`.
This method is SuSiE-relax-prime. The single-effect prime model
uses a latent relaxed contribution vector instead of directly sampling a relaxed
LD column in the likelihood:
\[
x \mid \theta_1,\ldots,\theta_L
\sim
N_J\left(
\sum_{\ell=1}^L \theta_\ell,
\frac{\bar R}{N}
\right).
\]
For each effect,
\[
\gamma_\ell \sim \operatorname{Unif}\{1,\ldots,J\},
\qquad
\beta_\ell \sim N(0,\sigma_\ell^2),
\]
and, conditional on \(\gamma_\ell=j\),
\[
\theta_\ell = \beta_\ell\bar R_{\cdot j}+\eta_{\ell j},
\qquad
\eta_{\ell j,j}=0,
\qquad
\eta_{\ell j,-j}\sim N_{J-1}(0,\sigma_\ell^2\tau^2S_j).
\]
Here
\[
\mu_j=\bar R_{-j,j},
\qquad
S_j=\bar R_{-j,-j}-\bar R_{-j,j}\bar R_{j,-j},
\qquad
\tau^2=(\nu_0+3)^{-1}.
\]

The variational approximation factorizes over effects:
\[
q =
\prod_{\ell=1}^L
q_\ell(\gamma_\ell,\beta_\ell,\eta_\ell).
\]
The inference algorithm is IBSS-like. When updating effect \(\ell\), we form the residual
\[
r_\ell = x-\sum_{k\ne\ell}\bar\theta_k,
\]
where
\[
\bar\theta_\ell =
\sum_{j=1}^J
\alpha_{\ell j}
\left[
m_{\ell j}\bar R_{\cdot j}
\;+\;
\rho_\ell(r_\ell-r_{\ell j}\bar R_{\cdot j})
\right].
\]
Then we apply the SER-relax-prime update to this residual. After each full sweep
over effects, the shared \(\nu_0\) is updated by the closed-form empirical-Bayes update
\[
\hat\tau^2 =
\frac{
\sum_{\ell=1}^L D_\ell/\sigma_\ell^2
}{L(J-1)},
\qquad
\hat\nu_0 = \frac{1}{\hat\tau^2}-3,
\]
with lower and upper bounds in implementation.

The main empirical behavior we want to demonstrate is:

- When the reference LD mismatch is low, ordinary SuSiE and SuSiE-relax-prime both recover the two true signals.

- When the reference LD mismatch is high, ordinary SuSiE with out-of-sample LD can overfit and select all \(L=5\) effects, while SuSiE-relax-prime can respond by learning a small \(\nu_0\), using stronger LD relaxation, and selecting fewer effects.

```{r helper-functions}
standardize_sim <- function(res) {
  d <- 1 / sqrt(diag(res$R))
  res$v <- d * res$v
  res$R <- cov2cor(res$R)
  res$R0 <- cov2cor(res$R0)
  res
}

make_rbar <- function(R0, lambda = 0.01) {
  Rbar <- (1 - lambda) * R0 + lambda * diag(1, ncol(R0))
  cov2cor(Rbar)
}

make_cs_from_alpha <- function(alpha, R, coverage = 0.95,
                               min_abs_corr = 0.5,
                               min_top_alpha = 1e-4) {
  alpha <- as.matrix(alpha)
  cs <- list()
  purity <- numeric()

  for (ell in seq_len(nrow(alpha))) {
    a <- as.numeric(alpha[ell, ])
    if (max(a) < min_top_alpha) {
      next
    }

    ord <- order(a, decreasing = TRUE)
    k <- which(cumsum(a[ord]) >= coverage)[1]
    vars <- ord[seq_len(k)]
    p <- if (length(vars) == 1) {
      1
    } else {
      R_sub <- abs(R[vars, vars, drop = FALSE])
      min(R_sub[upper.tri(R_sub)])
    }

    if (is.finite(p) && p >= min_abs_corr) {
      nm <- paste0("L", ell)
      cs[[nm]] <- vars
      purity[nm] <- p
    }
  }

  list(cs = cs, purity = purity)
}

count_selected_cs <- function(alpha, R, coverage = 0.95,
                              min_abs_corr = 0.5,
                              min_top_alpha = 1e-4) {
  length(make_cs_from_alpha(
    alpha = alpha,
    R = R,
    coverage = coverage,
    min_abs_corr = min_abs_corr,
    min_top_alpha = min_top_alpha
  )$cs)
}

summarize_fit <- function(label, alpha, pip, R_for_cs, beta_true,
                          nu0 = NA_real_) {
  alpha <- as.matrix(alpha)
  top <- apply(alpha, 1, which.max)
  top_alpha <- apply(alpha, 1, max)
  cs_info <- make_cs_from_alpha(alpha, R = R_for_cs)

  data.frame(
    method = label,
    selected_L = length(cs_info$cs),
    nu0 = nu0,
    gamma_hat = paste(top, collapse = ", "),
    top_alpha = paste(sprintf("%.3f", top_alpha), collapse = ", "),
    causal_pip = paste(sprintf("%.3f", pip[beta_true != 0]), collapse = ", ")
  )
}

plot_pip <- function(fit, R_for_cs, beta_true, main) {
  susie_plot_alpha(
    fit,
    R = R_for_cs,
    b = beta_true,
    main = main,
    coverage = 0.95,
    min_abs_corr = 0.5
  )
  grid()
}

plot_convergence <- function(fit, main_prefix) {
  old_par <- par(no.readonly = TRUE)
  on.exit(par(old_par), add = TRUE)

  par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
  elbo_iter <- which(!is.na(fit$elbo))
  plot(elbo_iter, fit$elbo[elbo_iter], type = "b", lwd = 2, pch = 16,
       xlab = "Iteration", ylab = "ELBO",
       main = paste(main_prefix, "ELBO"))
  grid()
  plot(seq_along(fit$nu0_trace), fit$nu0_trace, type = "l", lwd = 2,
       xlab = "Iteration", ylab = expression(hat(nu)[0]),
       main = paste(main_prefix, expression(nu[0])))
  grid()
}

plot_tau2_convergence <- function(fit, main_prefix) {
  tau2_trace <- 1 / (fit$nu0_trace + 3)
  plot(
    seq_along(tau2_trace), tau2_trace,
    type = "l",
    lwd = 2,
    xlab = "Iteration",
    ylab = expression(1 / (hat(nu)[0] + 3)),
    main = paste(main_prefix, expression(1 / (nu[0] + 3)))
  )
  grid()
}

run_case <- function(case_label, N0, seed, N = 20000L, J = 1000L,
                     lambda = 0.01, L = 5,
                     prime_max_iter = 3000,
                     prime_nu0_init = 2000,
                     prime_nu0_bounds = c(10, 10000),
                     prime_elbo_update_interval = 10) {
  res <- sim_2pop(
    chrom = "chr22",
    start = 20e6,
    end = 21e6,
    T_split = 20,
    n_gwas = N,
    n_ref = N0,
    J = J,
    h2 = 0.01,
    seed = seed
  )
  res <- standardize_sim(res)

  v <- res$v
  R <- res$R
  R0 <- res$R0
  Rbar <- make_rbar(R0, lambda = lambda)
  beta_true <- res$beta_true

  fit_susie_out <- susie_suff_stat(
    XtX = N * R0,
    Xty = N * v,
    n = N,
    yty = N,
    L = L
  )

  prime_fit_time <- system.time({
    fit_prime <- fit_susie_relax_prime(
      x = v,
      Rbar = Rbar,
      N = N,
      L = L,
      sigma2 = rep(0.3^2, L),
      nu0_init = prime_nu0_init,
      estimate_nu0 = TRUE,
      warmup_iter = 5,
      max_iter = prime_max_iter,
      tol = 1e-7,
      verbose = FALSE,
      nu0_bounds = prime_nu0_bounds,
      elbo_update_interval = prime_elbo_update_interval,
      nu0_update_interval = 1,
      nu0_tol = 1e-4
    )
  })

  summary <- rbind(
    summarize_fit(
      label = "susieR with R0",
      alpha = fit_susie_out$alpha,
      pip = fit_susie_out$pip,
      R_for_cs = R0,
      beta_true = beta_true
    ),
    summarize_fit(
      label = "SuSiE-relax-prime",
      alpha = fit_prime$alpha,
      pip = fit_prime$pip,
      R_for_cs = Rbar,
      beta_true = beta_true,
      nu0 = fit_prime$nu0
    )
  )

  list(
    case_label = case_label,
    N = N,
    N0 = N0,
    seed = seed,
    lambda = lambda,
    L = L,
    v = v,
    R = R,
    R0 = R0,
    Rbar = Rbar,
    beta_true = beta_true,
    causal = which(beta_true != 0),
    fit_susie_out = fit_susie_out,
    fit_prime = fit_prime,
    prime_fit_time = prime_fit_time,
    summary = summary
  )
}
```

# Low-mismatch Case

Here the reference panel size is \(N_0=1000\), so the out-of-sample LD is close enough to the GWAS LD. We expect both ordinary SuSiE and SuSiE-relax-prime to select two effects.

```{r low-mismatch-fit}
low_case <- run_case(
  case_label = "low mismatch",
  N0 = 2000L,
  seed = 6,
  N = 20000L,
  J = 1000L,
  lambda = 0.01,
  L = 5,
  prime_max_iter = 1000,
  prime_nu0_init = 2000,
  prime_nu0_bounds = c(10, 10000),
  prime_elbo_update_interval = 5
)

low_case$prime_fit_time
low_case$causal
low_case$summary
```

```{r low-mismatch-plots}
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
plot_pip(
  low_case$fit_susie_out,
  low_case$R0,
  low_case$beta_true,
  main = "Low mismatch: susieR with R0"
)
plot_pip(
  low_case$fit_prime,
  low_case$Rbar,
  low_case$beta_true,
  main = "Low mismatch: SuSiE-relax-prime"
)
par(old_par)

plot_convergence(low_case$fit_prime, "Low mismatch SuSiE-relax-prime")
```

Printing the ELBO (only compute once every few iterations because this is the most expensive step)

```{r}
low_case$fit_prime$elbo
```


It might make more sense to look at \(1/(\nu_0+3)\) instead of \(\nu_0\), because this is the relaxed-column variance scale used by the model. Large changes in \(\nu_0\) at high values may correspond to very small changes in the actual relaxation variance.

```{r low-mismatch-tau2-convergence}
plot_tau2_convergence(low_case$fit_prime, "Low mismatch SuSiE-relax-prime")
```

# High-mismatch Case (due to small N0)

Here the reference panel size is \(N_0=100\). The out-of-sample LD is much noisier. We fit ordinary SuSiE with upper bound \(L=5\) using the out-of-sample LD \(R_0\), and fit SuSiE-relax-prime with the stabilized matrix
\[
\bar R = 0.99R_0 + 0.01I.
\]
The expected behavior is that ordinary SuSiE overfits and selects all five effects, while SuSiE-relax-prime learns a small \(\nu_0\), relaxes the latent contribution, and selects fewer effects.

```{r high-mismatch-fit}
high_case <- run_case(
  case_label = "high mismatch",
  N0 = 100L,
  seed = 9,
  N = 20000L,
  J = 1000L,
  lambda = 0.01,
  L = 5,
  prime_max_iter = 3000,
  prime_nu0_init = 2000,
  prime_nu0_bounds = c(10, 10000),
  prime_elbo_update_interval = 5
)

high_case$prime_fit_time
high_case$causal
high_case$summary
```

```{r high-mismatch-plots}
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
plot_pip(
  high_case$fit_susie_out,
  high_case$R0,
  high_case$beta_true,
  main = "High mismatch: susieR with R0"
)
plot_pip(
  high_case$fit_prime,
  high_case$Rbar,
  high_case$beta_true,
  main = "High mismatch: SuSiE-relax-prime"
)
par(old_par)

plot_convergence(high_case$fit_prime, "High mismatch SuSiE-relax-prime")
```

```{r high-mismatch-effect-details}
data.frame(
  effect = seq_len(nrow(high_case$fit_prime$alpha)),
  top = apply(high_case$fit_prime$alpha, 1, which.max),
  top_alpha = apply(high_case$fit_prime$alpha, 1, max)
)
```


# High-mismatch Case (due to difference ancestry)

```{r}
sim_asn_eur <- function(chrom, start, end, n_gwas, n_ref, J, h2, seed) {

  stdpopsim <- reticulate::import("stdpopsim")
  msprime   <- reticulate::import("msprime")

  set.seed(seed)

  # 1. Get Realistic Genetic Map from stdpopsim
  species <- stdpopsim$get_species("HomSap")
  contig  <- species$get_contig(chrom, left = as.integer(start), right = as.integer(end))

  # 2. Get OutOfAfrica_3G09 Demography (YRI, CEU, CHB)
  demo_model <- species$get_demographic_model("OutOfAfrica_3G09")
  demography <- demo_model$model

  # 3. Run Simulation
  ts <- msprime$sim_ancestry(
    samples = reticulate::dict(CHB = as.integer(n_gwas), CEU = as.integer(n_ref)),
    demography = demography,
    recombination_rate = contig$recombination_map,
    random_seed = as.integer(seed)
  )

  ts <- msprime$sim_mutations(ts, rate = 1.29e-8, random_seed = as.integer(seed))

  # 4. Extract Genotypes
  chb_id <- NULL
  ceu_id <- NULL
  for (p in reticulate::iterate(ts$populations())) {
    if (!is.null(p$metadata) && p$metadata$name == "CHB") chb_id <- p$id
    if (!is.null(p$metadata) && p$metadata$name == "CEU") ceu_id <- p$id
  }

  samples_gwas <- ts$samples(population = as.integer(chb_id))
  samples_ref  <- ts$samples(population = as.integer(ceu_id))

  G_all <- ts$genotype_matrix()

  to_diploid <- function(hap_idx) {
    idx <- as.integer(hap_idx) + 1
    g_hap <- G_all[, idx, drop = FALSE]
    g_dip <- g_hap[, seq(1, ncol(g_hap), by = 2)] + g_hap[, seq(2, ncol(g_hap), by = 2)]
    return(t(g_dip))
  }

  G_gwas <- to_diploid(samples_gwas)
  G_ref  <- to_diploid(samples_ref)

  # 5. Filter for Common Variants (MAF > 5%)
  maf_gwas <- colMeans(G_gwas) / 2
  maf_ref  <- colMeans(G_ref) / 2

  common_mask <- (maf_gwas > 0.05) & (maf_gwas < 0.95) & (maf_ref > 0.05) & (maf_ref < 0.95)
  n_common <- sum(common_mask)
  if (n_common < J) stop(paste0("Only ", n_common, " common variants found but J = ", J, ". Try a larger region."))
  G_gwas <- G_gwas[, common_mask, drop = FALSE]
  G_ref  <- G_ref[, common_mask, drop = FALSE]

  J_all <- min(ncol(G_gwas), J)
  max_start <- J_all - J
  start_idx <- if (max_start > 0) sample(0:max_start, 1) else 0
  cols_to_keep <- (start_idx + 1):(start_idx + J)
  G  <- G_gwas[, cols_to_keep, drop = FALSE]
  G0 <- G_ref[, cols_to_keep, drop = FALSE]

  # 6. Generate summary stats
  X <- sweep(G, 2, colMeans(G), "-")
  N <- nrow(X)

  G0_centered <- sweep(G0, 2, colMeans(G0), "-")

  R0 <- stats::cov(G0_centered)
  R0[is.na(R0)] <- 0

  R <- stats::cov(X)
  R[is.na(R)] <- 0

  num_causal <- 2
  first_causal_SNP <- sample(1:J, 1)
  second_causal_SNP <- which.min(R[first_causal_SNP, ]^2)
  causal_SNPs_loc <- c(first_causal_SNP, second_causal_SNP)

  beta_true <- rep(0, J)
  beta_true[causal_SNPs_loc] <- stats::runif(num_causal) + 0.01

  var_expl <- sum((X %*% beta_true)^2 / N)
  scale <- h2 / var_expl
  beta_true <- beta_true * sqrt(scale)

  y <- (X %*% beta_true) + sqrt(1 - h2) * stats::rnorm(N)
  y_std <- sqrt(sum((y - mean(y))^2) / N)
  y <- (y - mean(y)) / y_std

  v <- (t(X) %*% y) / N

  return(list(
    v = as.vector(v),
    R = R,
    R0 = R0,
    beta_true = beta_true,
    causal_SNPs_loc = causal_SNPs_loc
  ))
}

```

```{r}
seed = 2
chrom <- "chr22"
start <- 20e6
end   <- 21e6
N     <- 25000L
N0 = 500L
J     <- 500L
h2    <- 0.01

res <- sim_asn_eur(chrom, start, end,
                n_gwas = N,
                n_ref = N0,
                J = J,
                h2 = h2,
                seed = seed)

v               <- res$v
R               <- res$R
R0              <- res$R0
beta_true       <- res$beta_true
causal_SNPs_loc <- res$causal_SNPs_loc

## standardized
d = 1 / sqrt(diag(R))
v = d * v
R = cov2cor(R)
R0 = cov2cor(R0)

x = v
plot(x, pch = 16)
points(causal_SNPs_loc, x[causal_SNPs_loc], pch = 16, col = "red")
```

```{r ancestry-mismatch-fit-prime}
lambda <- 0.1
L <- 5
Rbar <- make_rbar(R0, lambda = lambda)

fit_susie_out_asn_eur <- susie_suff_stat(
  XtX = N * R0,
  Xty = N * x,
  n = N,
  yty = N,
  L = L
)

prime_fit_time_asn_eur <- system.time({
  fit_prime_asn_eur <- fit_susie_relax_prime(
    x = x,
    Rbar = Rbar,
    N = N,
    L = L,
    sigma2 = rep(0.3^2, L),
    nu0_init = 2000,
    estimate_nu0 = TRUE,
    warmup_iter = 5,
    max_iter = 3000,
    tol = 1e-7,
    verbose = FALSE,
    nu0_bounds = c(10, 10000),
    elbo_update_interval = 5,
    nu0_update_interval = 1,
    nu0_tol = 1e-4
  )
})

prime_fit_time_asn_eur
causal_SNPs_loc
rbind(
  summarize_fit(
    label = "susieR with R0",
    alpha = fit_susie_out_asn_eur$alpha,
    pip = fit_susie_out_asn_eur$pip,
    R_for_cs = R0,
    beta_true = beta_true
  ),
  summarize_fit(
    label = "SuSiE-relax-prime",
    alpha = fit_prime_asn_eur$alpha,
    pip = fit_prime_asn_eur$pip,
    R_for_cs = Rbar,
    beta_true = beta_true,
    nu0 = fit_prime_asn_eur$nu0
  )
)
```

```{r ancestry-mismatch-prime-plots}
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
plot_pip(
  fit_susie_out_asn_eur,
  R0,
  beta_true,
  main = "Ancestry mismatch: susieR with R0"
)
plot_pip(
  fit_prime_asn_eur,
  Rbar,
  beta_true,
  main = "Ancestry mismatch: SuSiE-relax-prime"
)
par(old_par)

plot_convergence(
  fit_prime_asn_eur,
  "Ancestry mismatch SuSiE-relax-prime"
)
```

```{r ancestry-mismatch-prime-effect-details}
data.frame(
  effect = seq_len(nrow(fit_prime_asn_eur$alpha)),
  top = apply(fit_prime_asn_eur$alpha, 1, which.max),
  top_alpha = apply(fit_prime_asn_eur$alpha, 1, max)
)
```
