---
title: "SuSiE-collapse-R Shared-Omega Analysis"
author: "Dat Do"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)

library(susieR)

source("code/SuSiE-collapse-R-shared-omega.R")
source("code/sim_2pop.R")
source("code/plot.R")
```

# Goal

This page repeats the main analysis in
`analysis/susie-relax-demonstrate.Rmd`, replacing `fit_susie_relax()`
with the collapsed-\(R\) SuSiE fit using one shared \(q(\omega)\):
`fit_susie_collapse_r_shared()`.

The comparison is between:

- ordinary `susieR` using the out-of-sample LD matrix \(R_0\);
- SuSiE-collapse-\(R\) using the stabilized reference matrix
  \(\bar R=(1-\lambda)R_0+\lambda I\).

Unlike `fit_susie_relax()`, the collapsed-\(R\) implementation estimates
\(\nu_0\) by profiling over a grid of candidate values.

```{r helper-functions}
standardize_sim <- function(res) {
  d <- 1 / sqrt(diag(res$R))
  res$v <- d * res$v
  res$R <- cov2cor(res$R)
  res$R0 <- cov2cor(res$R0)
  res
}

make_rbar <- function(R0, lambda = 0.01) {
  Rbar <- (1 - lambda) * R0 + lambda * diag(1, ncol(R0))
  cov2cor(Rbar)
}

make_cs_from_alpha <- function(alpha, R, coverage = 0.95,
                               min_abs_corr = 0.5,
                               min_top_alpha = 1e-4) {
  alpha <- as.matrix(alpha)
  cs <- list()
  purity <- numeric()

  for (ell in seq_len(nrow(alpha))) {
    a <- as.numeric(alpha[ell, ])
    if (max(a) < min_top_alpha) {
      next
    }

    ord <- order(a, decreasing = TRUE)
    k <- which(cumsum(a[ord]) >= coverage)[1]
    vars <- ord[seq_len(k)]
    p <- if (length(vars) == 1) {
      1
    } else {
      R_sub <- abs(R[vars, vars, drop = FALSE])
      min(R_sub[upper.tri(R_sub)])
    }

    if (is.finite(p) && p >= min_abs_corr) {
      nm <- paste0("L", ell)
      cs[[nm]] <- vars
      purity[nm] <- p
    }
  }

  list(cs = cs, purity = purity)
}

summarize_fit <- function(label, alpha, pip, R_for_cs, beta_true,
                          nu0 = NA_real_, sigma2 = NA_real_) {
  alpha <- as.matrix(alpha)
  top <- apply(alpha, 1, which.max)
  top_alpha <- apply(alpha, 1, max)
  cs_info <- make_cs_from_alpha(alpha, R = R_for_cs)

  data.frame(
    method = label,
    selected_L = length(cs_info$cs),
    nu0 = nu0,
    sigma2 = paste(sprintf("%.4g", sigma2), collapse = ", "),
    gamma_hat = paste(top, collapse = ", "),
    top_alpha = paste(sprintf("%.3f", top_alpha), collapse = ", "),
    causal_pip = paste(sprintf("%.3f", pip[beta_true != 0]), collapse = ", ")
  )
}

plot_pip <- function(fit, R_for_cs, beta_true, main) {
  susie_plot_alpha(
    fit,
    R = R_for_cs,
    b = beta_true,
    main = main,
    coverage = 0.95,
    min_abs_corr = 0.5
  )
  grid()
}

plot_collapse_convergence <- function(fit, main_prefix) {
  old_par <- par(no.readonly = TRUE)
  on.exit(par(old_par), add = TRUE)

  par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
  plot(
    seq_along(fit$elbo), fit$elbo,
    type = "b",
    pch = 16,
    xlab = "iteration",
    ylab = "lower bound",
    main = paste(main_prefix, "best-grid fit")
  )
  grid()
  plot(
    fit$nu0_grid, fit$grid_lower_bound,
    type = "b",
    pch = 16,
    xlab = expression(nu[0]),
    ylab = "optimized lower bound",
    main = paste(main_prefix, expression(nu[0] * " profile"))
  )
  abline(v = fit$nu0, col = 2, lty = 2)
  grid()
}

run_case <- function(case_label, N0, seed, N = 20000L, J = 1000L,
                     lambda = 0.01, L = 5,
                     nu0_grid = c(10, 25, 50, 100, 200, 500, 1000, 2000, 5000),
                     collapse_max_iter = 200,
                     collapse_sigma2 = 0.3^2,
                     estimate_sigma2 = TRUE) {
  res <- sim_2pop(
    chrom = "chr22",
    start = 20e6,
    end = 21e6,
    T_split = 20,
    n_gwas = N,
    n_ref = N0,
    J = J,
    h2 = 0.01,
    seed = seed
  )
  res <- standardize_sim(res)

  v <- res$v
  R <- res$R
  R0 <- res$R0
  Rbar <- make_rbar(R0, lambda = lambda)
  beta_true <- res$beta_true

  fit_susie_out <- susie_suff_stat(
    XtX = N * R0,
    Xty = N * v,
    n = N,
    yty = N,
    L = L
  )

  collapse_fit_time <- system.time({
    fit_collapse <- fit_susie_collapse_r_shared(
      x = v,
      Rbar = Rbar,
      N = N,
      L = L,
      sigma2 = rep(collapse_sigma2, L),
      nu0_grid = nu0_grid,
      estimate_sigma2 = estimate_sigma2,
      max_iter = collapse_max_iter,
      tol = 1e-6,
      verbose = FALSE
    )
  })

  summary <- rbind(
    summarize_fit(
      label = "susieR with R0",
      alpha = fit_susie_out$alpha,
      pip = fit_susie_out$pip,
      R_for_cs = R0,
      beta_true = beta_true
    ),
    summarize_fit(
      label = "SuSiE-collapse-R shared q(omega)",
      alpha = fit_collapse$alpha,
      pip = fit_collapse$pip,
      R_for_cs = Rbar,
      beta_true = beta_true,
      nu0 = fit_collapse$nu0,
      sigma2 = fit_collapse$sigma2
    )
  )

  list(
    case_label = case_label,
    N = N,
    N0 = N0,
    seed = seed,
    lambda = lambda,
    L = L,
    v = v,
    R = R,
    R0 = R0,
    Rbar = Rbar,
    beta_true = beta_true,
    causal = which(beta_true != 0),
    fit_susie_out = fit_susie_out,
    fit_collapse = fit_collapse,
    collapse_fit_time = collapse_fit_time,
    summary = summary
  )
}
```

# Low-mismatch Case

Here the reference panel is large. As in the original SuSiE-relax
analysis, ordinary SuSiE and the relaxed method should both have an
easier problem.

```{r low-mismatch-fit}
low_case <- run_case(
  case_label = "low mismatch",
  N0 = 2000L,
  seed = 6,
  N = 20000L,
  J = 1000L,
  lambda = 0.01,
  L = 5,
  collapse_max_iter = 200,
  collapse_sigma2 = 0.3^2
)

low_case$collapse_fit_time
low_case$causal
low_case$summary
```

```{r low-mismatch-plots}
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
plot_pip(
  low_case$fit_susie_out,
  low_case$R0,
  low_case$beta_true,
  main = "Low mismatch: susieR with R0"
)
plot_pip(
  low_case$fit_collapse,
  low_case$Rbar,
  low_case$beta_true,
  main = "Low mismatch: SuSiE-collapse-R"
)
par(old_par)

plot_collapse_convergence(low_case$fit_collapse, "Low mismatch")
```

# High-mismatch Case

Here the reference panel is small. This repeats the high-mismatch setup
from the original page, but fits the collapsed-\(R\) shared-\(\omega\)
model.

```{r high-mismatch-fit}
high_case <- run_case(
  case_label = "high mismatch",
  N0 = 100L,
  seed = 9,
  N = 20000L,
  J = 1000L,
  lambda = 0.01,
  L = 5,
  collapse_max_iter = 200,
  collapse_sigma2 = 0.3^2
)

high_case$collapse_fit_time
high_case$causal
high_case$summary
```

```{r high-mismatch-plots}
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
plot_pip(
  high_case$fit_susie_out,
  high_case$R0,
  high_case$beta_true,
  main = "High mismatch: susieR with R0"
)
plot_pip(
  high_case$fit_collapse,
  high_case$Rbar,
  high_case$beta_true,
  main = "High mismatch: SuSiE-collapse-R"
)
par(old_par)

plot_collapse_convergence(high_case$fit_collapse, "High mismatch")
```

```{r high-mismatch-effect-details}
data.frame(
  effect = seq_len(nrow(high_case$fit_collapse$alpha)),
  top = apply(high_case$fit_collapse$alpha, 1, which.max),
  top_alpha = apply(high_case$fit_collapse$alpha, 1, max)
)
```

# High-mismatch Case Due To Different Ancestry

```{r ancestry-mismatch-simulation-function}
sim_asn_eur <- function(chrom, start, end, n_gwas, n_ref, J, h2, seed) {

  stdpopsim <- reticulate::import("stdpopsim")
  msprime   <- reticulate::import("msprime")

  set.seed(seed)

  species <- stdpopsim$get_species("HomSap")
  contig  <- species$get_contig(chrom, left = as.integer(start), right = as.integer(end))

  demo_model <- species$get_demographic_model("OutOfAfrica_3G09")
  demography <- demo_model$model

  ts <- msprime$sim_ancestry(
    samples = reticulate::dict(CHB = as.integer(n_gwas), CEU = as.integer(n_ref)),
    demography = demography,
    recombination_rate = contig$recombination_map,
    random_seed = as.integer(seed)
  )

  ts <- msprime$sim_mutations(ts, rate = 1.29e-8, random_seed = as.integer(seed))

  chb_id <- NULL
  ceu_id <- NULL
  for (p in reticulate::iterate(ts$populations())) {
    if (!is.null(p$metadata) && p$metadata$name == "CHB") chb_id <- p$id
    if (!is.null(p$metadata) && p$metadata$name == "CEU") ceu_id <- p$id
  }

  samples_gwas <- ts$samples(population = as.integer(chb_id))
  samples_ref  <- ts$samples(population = as.integer(ceu_id))

  G_all <- ts$genotype_matrix()

  to_diploid <- function(hap_idx) {
    idx <- as.integer(hap_idx) + 1
    g_hap <- G_all[, idx, drop = FALSE]
    g_dip <- g_hap[, seq(1, ncol(g_hap), by = 2)] +
      g_hap[, seq(2, ncol(g_hap), by = 2)]
    t(g_dip)
  }

  G_gwas <- to_diploid(samples_gwas)
  G_ref  <- to_diploid(samples_ref)

  maf_gwas <- colMeans(G_gwas) / 2
  maf_ref  <- colMeans(G_ref) / 2

  common_mask <- (maf_gwas > 0.05) & (maf_gwas < 0.95) &
    (maf_ref > 0.05) & (maf_ref < 0.95)
  n_common <- sum(common_mask)
  if (n_common < J) {
    stop(paste0("Only ", n_common, " common variants found but J = ", J,
                ". Try a larger region."))
  }
  G_gwas <- G_gwas[, common_mask, drop = FALSE]
  G_ref  <- G_ref[, common_mask, drop = FALSE]

  J_all <- min(ncol(G_gwas), J)
  max_start <- J_all - J
  start_idx <- if (max_start > 0) sample(0:max_start, 1) else 0
  cols_to_keep <- (start_idx + 1):(start_idx + J)
  G  <- G_gwas[, cols_to_keep, drop = FALSE]
  G0 <- G_ref[, cols_to_keep, drop = FALSE]

  X <- sweep(G, 2, colMeans(G), "-")
  N <- nrow(X)

  G0_centered <- sweep(G0, 2, colMeans(G0), "-")

  R0 <- stats::cov(G0_centered)
  R0[is.na(R0)] <- 0

  R <- stats::cov(X)
  R[is.na(R)] <- 0

  num_causal <- 2
  first_causal_SNP <- sample(1:J, 1)
  second_causal_SNP <- which.min(R[first_causal_SNP, ]^2)
  causal_SNPs_loc <- c(first_causal_SNP, second_causal_SNP)

  beta_true <- rep(0, J)
  beta_true[causal_SNPs_loc] <- stats::runif(num_causal) + 0.01

  var_expl <- sum((X %*% beta_true)^2 / N)
  scale <- h2 / var_expl
  beta_true <- beta_true * sqrt(scale)

  y <- (X %*% beta_true) + sqrt(1 - h2) * stats::rnorm(N)
  y_std <- sqrt(sum((y - mean(y))^2) / N)
  y <- (y - mean(y)) / y_std

  v <- (t(X) %*% y) / N

  list(
    v = as.vector(v),
    R = R,
    R0 = R0,
    beta_true = beta_true,
    causal_SNPs_loc = causal_SNPs_loc
  )
}
```

```{r ancestry-mismatch-fit}
seed <- 4
chrom <- "chr22"
start <- 20e6
end <- 21e6
N <- 25000L
N0 <- 500L
J <- 500L
h2 <- 0.01

res <- sim_asn_eur(
  chrom, start, end,
  n_gwas = N,
  n_ref = N0,
  J = J,
  h2 = h2,
  seed = seed
)

v <- res$v
R <- res$R
R0 <- res$R0
beta_true <- res$beta_true
causal_SNPs_loc <- res$causal_SNPs_loc

d <- 1 / sqrt(diag(R))
v <- d * v
R <- cov2cor(R)
R0 <- cov2cor(R0)
x <- v

L <- 5
lambda_grid <- c(1e-4, 1e-3, 1e-2, 0.05, 0.1, 0.15, 0.2, 0.3, 0.4, 0.5)
nu0_grid <- c(10, 25, 50, 100, 200, 500, 1000, 2000, 5000)

fit_susie_out_asn_eur <- susie_suff_stat(
  XtX = N * R0,
  Xty = N * x,
  n = N,
  yty = N,
  L = L
)

collapse_fit_time_asn_eur <- system.time({
  collapse_lambda_fits <- lapply(lambda_grid, function(lambda) {
    fit_susie_collapse_r_shared(
      x = x,
      Rbar = make_rbar(R0, lambda = lambda),
      N = N,
      L = L,
      sigma2 = rep(0.3^2, L),
      nu0_grid = nu0_grid,
      estimate_sigma2 = TRUE,
      max_iter = 200,
      tol = 1e-6,
      verbose = FALSE
    )
  })
})
names(collapse_lambda_fits) <- sprintf("lambda=%g", lambda_grid)

lambda_summary <- do.call(rbind, lapply(seq_along(lambda_grid), function(i) {
  fit <- collapse_lambda_fits[[i]]
  Rbar_i <- make_rbar(R0, lambda = lambda_grid[i])
  data.frame(
    lambda = lambda_grid[i],
    final_elbo = fit$lower_bound,
    nu0 = fit$nu0,
    selected_L = length(make_cs_from_alpha(fit$alpha, R = Rbar_i)$cs),
    causal_pip = paste(sprintf("%.3f", fit$pip[beta_true != 0]), collapse = ", "),
    top = paste(apply(fit$alpha, 1, which.max), collapse = ", "),
    top_alpha = paste(sprintf("%.3f", apply(fit$alpha, 1, max)), collapse = ", ")
  )
}))

best_idx <- which.max(lambda_summary$final_elbo)
best_lambda <- lambda_summary$lambda[best_idx]
Rbar <- make_rbar(R0, lambda = best_lambda)
fit_collapse_asn_eur <- collapse_lambda_fits[[best_idx]]

collapse_fit_time_asn_eur
causal_SNPs_loc
lambda_summary
best_lambda
rbind(
  summarize_fit(
    label = "susieR with R0",
    alpha = fit_susie_out_asn_eur$alpha,
    pip = fit_susie_out_asn_eur$pip,
    R_for_cs = R0,
    beta_true = beta_true
  ),
  summarize_fit(
    label = sprintf("SuSiE-collapse-R shared q(omega), lambda = %g", best_lambda),
    alpha = fit_collapse_asn_eur$alpha,
    pip = fit_collapse_asn_eur$pip,
    R_for_cs = Rbar,
    beta_true = beta_true,
    nu0 = fit_collapse_asn_eur$nu0,
    sigma2 = fit_collapse_asn_eur$sigma2
  )
)
```

```{r ancestry-mismatch-lambda-profile}
plot(
  lambda_summary$lambda,
  lambda_summary$final_elbo,
  type = "b",
  pch = 16,
  log = "x",
  xlab = expression(lambda),
  ylab = "optimized lower bound",
  main = "Ancestry mismatch: lower bound vs lambda"
)
points(
  best_lambda,
  lambda_summary$final_elbo[best_idx],
  col = 2,
  pch = 19,
  cex = 1.4
)
grid()
```

```{r ancestry-mismatch-plots}
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
plot_pip(
  fit_susie_out_asn_eur,
  R0,
  beta_true,
  main = "Ancestry mismatch: susieR with R0"
)
plot_pip(
  fit_collapse_asn_eur,
  Rbar,
  beta_true,
  main = sprintf("Ancestry mismatch: SuSiE-collapse-R, lambda = %g", best_lambda)
)
par(old_par)

plot_collapse_convergence(
  fit_collapse_asn_eur,
  "Ancestry mismatch"
)
```

```{r ancestry-mismatch-effect-details}
data.frame(
  effect = seq_len(nrow(fit_collapse_asn_eur$alpha)),
  top = apply(fit_collapse_asn_eur$alpha, 1, which.max),
  top_alpha = apply(fit_collapse_asn_eur$alpha, 1, max)
)
```
